---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# fabricQueryR

<!-- badges: start -->
[![R-CMD-check](https://github.com/kennispunttwente/fabricQueryR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/kennispunttwente/fabricQueryR/actions/workflows/R-CMD-check.yaml)
[![CRAN status](https://www.r-pkg.org/badges/version/fabricQueryR)](https://CRAN.R-project.org/package=fabricQueryR)
<!-- badges: end -->

'fabricQueryR' is an R package which helps you discover and query data from Microsoft Fabric in R.
It comes with discovery helpers and five methods which help you to get your Microsoft Fabric data into R:

1. Create a connection to a SQL endpoint (e.g., from a `Lakehouse` or `Data Warehouse` item): `fabric_sql_connect()`.
This results in a 'DBI' connection object which you can execute SQL queries with, and/or
use with 'DBI'-compatible packages like 'dbplyr'.

2. Execute a DAX query against a Fabric/Power Bi `Semantic Model` item: `fabric_pbi_dax_query()`.
With this, you can run DAX queries against a Fabric/Power Bi dataset and get the results as a 'tibble' dataframe.

3. Read a Delta table from a Fabric `Lakehouse` item: `fabric_onelake_read_delta_table()`.
This function downloads the underlying Parquet files from the Delta table stored in OneLake (ADLS
Gen2) and returns the data as a 'tibble' dataframe.

4. Execute a Livy API query: `fabric_livy_query()`.
With this, you can remotely execute Spark/Spark SQL/SparkR/PySpark code in Microsoft Fabric
and get a list with the results in your local R session.

5. Execute a KQL query against an `Eventhouse`/`KQL Database` item:
`fabric_kql_query()`.
This returns typed Kusto query results as a tibble, or a named list of tibbles
when KQL returns multiple primary result tables.

## Installation

You can install the development version of 'fabricQueryR' like so:

```{r, eval=FALSE}
if (!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes")
}

remotes::install_github("kennispunttwente/fabricQueryR")
```

Or, install the latest release from CRAN:

```{r, eval=FALSE}
install.packages("fabricQueryR")
```

## Usage

See the [reference](https://kennispunttwente.github.io/fabricQueryR/reference/index.html)
for the full documentation of all functions.

Below is a code snippet showing how to discover targets and use the five methods to get data from Fabric into R:

```{r, eval=FALSE}

# First find your 'tenant' ID & 'client' ID (app registration) in Azure/Entra
# You may be able to use the default Azure CLI app id;
#   this will be automatically used if you do not set 'FABRICQUERYR_CLIENT_ID'
# The AzureAuth package is used to acquire tokens; you may be redirected
#   to a browser window to sign in the first time

library(fabricQueryR)

# Sys.setenv(FABRICQUERYR_TENANT_ID = "...")
# Sys.setenv(FABRICQUERYR_CLIENT_ID = "...")

# Discover workspaces and items ------------------------------------------------

workspaces <- fabric_workspaces()
workspace <- workspaces[workspaces$displayName == "ExampleWorkspace", ]

# Typed helpers retrieve workload-specific properties and ready-to-use targets.
lakehouse <- fabric_lakehouses(workspace)[1, ]
semantic_model <- fabric_semantic_models(workspace)[1, ]
kql_database <- fabric_kql_databases(workspace)[1, ]

# Other helpers include fabric_warehouses(), fabric_sql_databases(),
# fabric_eventhouses(), fabric_kql_databases(), fabric_notebooks(), and
# fabric_graphql_apis().


# SQL connection to Warehouse, Lakehouse, or SQL Database ----------------------

# A discovered item supplies both the server and required database/catalog.
# You can alternatively provide a complete portal connection string, or a bare
# endpoint together with database = "...".

# Get connection
con <- fabric_sql_connect(lakehouse)

# List databases
DBI::dbGetQuery(con, "SELECT name FROM sys.databases")

# List tables in the current database
DBI::dbGetQuery(
  con,
  "
  SELECT TABLE_SCHEMA, TABLE_NAME
  FROM INFORMATION_SCHEMA.TABLES
  WHERE TABLE_TYPE = 'BASE TABLE'
  "
)

# Read 'Customers' table
df_sql <- fabric_sql_query(
  lakehouse,
  "SELECT * FROM dbo.Customers WHERE region = ?",
  params = list("West")
)

# Close connection
DBI::dbDisconnect(con)


# Table from Lakehouse via OneLake data access ---------------------------------

# Ensure that the account/principal you authenticate with has access via
#   being part of the workspace, or via Lakehouse -> Manage OneLake data access

df_onelake <- fabric_onelake_read_delta_table(
  table_path = "Customers",
  workspace_name = workspace,
  lakehouse_name = lakehouse
)


# DAX query against Semantic Model ---------------------------------------------

# Ensure that the account you use to authenticate has access to the workspace,
#   or that you have been granted 'Build' permissions on the dataset (via share)

df_dax <- fabric_pbi_dax_query(
  connstr = semantic_model,
  dax = "EVALUATE TOPN(100000, 'Sheet1')"
)


# Livy API query to execute Spark code -----------------------------------------

# Ensure that the account you use to authenticate has access to the workspace

# Run a Livy SparkR query
livy_sparkr_result <- fabric_livy_query(
  livy_url = lakehouse,
  kind = "sparkr",
  code = "print(1+2)"
)

# (See example in `?fabric_livy_query` for how to get data from the Lakehouse to R with Spark)


# KQL query against an Eventhouse database ------------------------------------

# Query parameters are bound separately from the KQL query text.
df_kql <- fabric_kql_query(
  kql_database,
  query = paste(
    "declare query_parameters(selected_type:string);",
    "Events | where EventType == selected_type | take 100"
  ),
  parameters = list(selected_type = "Warning")
)

```

## Background

Microsoft Fabric is a new data platform from Microsoft which combines various data services,
including data warehousing, data lakes, and business intelligence.
It is built on top of Azure Data Services and integrates with Power BI for analytics and reporting.
Microsoft is actively promoting Fabric as the next-generation data platform for organizations using Microsoft Azure and Power BI.

As our organization started working with Microsoft Fabric, I found that that loading data into R from Fabric was not
yet straightforward, and took some effort to get working. To help others in the same situation,
I decided to share the functions I created to make this easier.
