---
title: "Authentication with Microsoft Fabric"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Authentication with Microsoft Fabric}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`fabricQueryR` uses
[`AzureAuth`](https://azure.r-universe.dev/AzureAuth/doc/token.html) for
Microsoft Entra authentication. Every exported function that authenticates
accepts the same four arguments:

- `tenant_id` and `client_id` identify the tenant and app registration. By default,
these are read from the `FABRICQUERYR_TENANT_ID` and `FABRICQUERYR_CLIENT_ID`
environment variables.
- `token` accepts an `AzureAuth::AzureToken` object, a bearer-token string, or
  a token-provider function. Its default is `NULL`, which triggers `AzureAuth` to acquire a token interactively or from its cache.
- `auth_args` is a named list of arguments forwarded to  `AzureAuth::get_azure_token()`.

When `token = NULL`, `fabricQueryR` calls `AzureAuth::get_azure_token()`.
`AzureAuth` first looks for a matching cached token and, if it cannot use one,
starts its normal interactive authorization-code or device-code login.
Use `auth_args` to select or configure that flow. When `token` is not `NULL`,
`auth_args` is not used. This package controls the token's `resource`, `tenant`, `app`,
and Azure AD `version` arguments, because those depend on the Fabric operation; all other
`get_azure_token()` arguments can go in `auth_args`.

## Set the tenant and client IDs

To tell `fabricQueryR` which Microsoft Entra tenant and app registration to use, set the
`FABRICQUERYR_TENANT_ID` and `FABRICQUERYR_CLIENT_ID` environment variables. 
You can do this in your `.Renviron` file, or in the R session:

```{r, eval = FALSE}
Sys.setenv(
  FABRICQUERYR_TENANT_ID = "<tenant-id>",
  FABRICQUERYR_CLIENT_ID = "<app-registration-client-id>"
)

library(fabricQueryR)

# Test by discovering workspaces; this will trigger an interactive login if no cached token is found
workspaces <- fabric_workspaces()
```

A tenant ID must always be set. In the
[Azure portal](https://portal.azure.com), open **Microsoft Entra ID >
Overview** and copy the **Tenant ID** shown under **Basic information**. This
is also called the *Directory (tenant) ID* in app-registration screens.

If `FABRICQUERYR_CLIENT_ID` is unset, the package uses the Azure CLI public
client ID (`04b07795-8ddb-461a-bbee-02f9e1bf7b46`), which may work in some tenants but not others.
A dedicated app registration is more reliable, and it can be scoped to a security group for least-privilege access.
Get in touch with your administrator to create an app registration if needed.

## Interactive login and the token cache

With `token = NULL`, `AzureAuth` first looks for a matching cached token. If
none exists, it chooses authorization-code login when a local browser flow is
available, otherwise device-code login. The first interactive login can
therefore open a browser or display a device code; later calls normally reuse
and refresh the cached token.

See the `AzureAuth` documentation on
[authentication scenarios](https://azure.r-universe.dev/AzureAuth/doc/scenarios.html)
and [caching](https://azure.r-universe.dev/AzureAuth/doc/token.html#caching).

Choose a flow explicitly when needed:

```{r, eval = FALSE}
# Remote R session: display a device code
items <- fabric_items(
  "Analytics",
  auth_args = list(auth_type = "device_code")
)

# Force a new login and do not read or write the AzureAuth cache
items <- fabric_items(
  "Analytics",
  auth_args = list(use_cache = FALSE)
)
```

Authorization-code login requires a suitable redirect URI on the app
registration (normally `http://localhost:1410`) and the `httpuv` package.
Device-code login requires **Allow public client flows** on the app
registration. Conditional Access and multifactor-authentication policies still
apply.

## Pass an AzureAuth token

You can acquire a token with `AzureAuth` and pass the resulting R6 object
directly. `fabricQueryR` extracts its access token, checks its expiry, and calls
its `refresh()` method when needed:

```{r, eval = FALSE}
fabric_token <- AzureAuth::get_azure_token(
  resource = c(
    "https://api.fabric.microsoft.com/.default",
    "offline_access"
  ),
  tenant = Sys.getenv("FABRICQUERYR_TENANT_ID"),
  app = Sys.getenv("FABRICQUERYR_CLIENT_ID"),
  version = 2
)

workspaces <- fabric_workspaces(token = fabric_token)
```

Tokens are resource-specific. Pass a token issued for the resource used by the
function:

| Operations | Azure AD v2 resource/scope |
|---|---|
| Discovery, Fabric REST, jobs, and Livy | `https://api.fabric.microsoft.com/.default` |
| OneLake files and Delta tables | `https://storage.azure.com/.default` |
| SQL connections | `https://database.windows.net/.default` |
| DAX / Power BI Execute Queries | `https://analysis.windows.net/powerbi/api/.default` |
| KQL | `https://api.kusto.windows.net/.default` |
| GraphQL as a delegated user | `https://analysis.windows.net/powerbi/api/GraphQLApi.Execute.All` |
| GraphQL as a service principal | `https://api.fabric.microsoft.com/.default` |

For a GraphQL service principal acquired by the package, the Fabric API
audience is selected automatically:

```{r, eval = FALSE}
result <- fabric_graphql_query(
  api,
  query,
  auth_args = list(
    password = Sys.getenv("FABRIC_CLIENT_SECRET"),
    auth_type = "client_credentials"
  )
)
```

Set `audience = "https://api.fabric.microsoft.com/.default"` explicitly when a
custom token-provider function obtains the service-principal token.

## Non-interactive authentication

Below are examples of non-interactive authentication flows. These avoid the
browser/device-code login and are suitable for CI/CD pipelines, scheduled jobs,
and service principals. They require a dedicated app registration and, for
service principals, a secret or certificate.

### Service principal with a client secret

`AzureAuth` calls this the `client_credentials` flow. Keep the secret outside
source code:

```{r, eval = FALSE}
workspaces <- fabric_workspaces(
  tenant_id = Sys.getenv("FABRICQUERYR_TENANT_ID"),
  client_id = Sys.getenv("FABRICQUERYR_CLIENT_ID"),
  auth_args = list(
    password = Sys.getenv("FABRIC_CLIENT_SECRET"),
    auth_type = "client_credentials"
  )
)
```

When client credentials are selected, the package does not request
`offline_access`, because application tokens do not use delegated refresh
tokens. `AzureAuth` reacquires them with the application credential.

### Service principal with a certificate

Certificate authentication avoids a long-lived client secret:

```{r, eval = FALSE}
workspaces <- fabric_workspaces(
  auth_args = list(
    certificate = Sys.getenv("FABRIC_CLIENT_CERTIFICATE"),
    auth_type = "client_credentials"
  )
)
```

`certificate` accepts the formats supported by `AzureAuth`, including a
PEM/PFX file and supported Azure Key Vault certificate objects. Protect the
certificate and its private key as credentials.

### Managed identity

Where the individual Fabric API supports managed identities, acquire an
`AzureToken` with `AzureAuth::get_managed_token()` and pass it through `token`:

```{r, eval = FALSE}
managed_token <- AzureAuth::get_managed_token(
  "https://api.fabric.microsoft.com"
)
workspaces <- fabric_workspaces(token = managed_token)
```

Managed-identity support is API-specific. Check the **Microsoft Entra supported
identities** table on the relevant Fabric REST API page. The current Fabric
[identity-support documentation](https://learn.microsoft.com/en-us/rest/api/fabric/articles/identity-support)
explains the tenant switch and this per-API requirement.

### Custom token brokers

For workload identity federation, an external secret store, or another token
broker, pass a function. It can accept `audience` and `force_refresh`:

```{r, eval = FALSE}
provider <- function(audience, force_refresh = FALSE) {
  # Acquire a token for `audience`; bypass your cache when force_refresh is TRUE.
  my_token_broker(audience, refresh = force_refresh)
}

files <- fabric_onelake_list(
  "Analytics",
  "Curated.Lakehouse",
  token = provider
)
```

The callback must return one bearer-token string (or a list containing an
`access_token` or `token` field). This is also the escape hatch for
authentication methods not implemented by `AzureAuth`.

## Configure authorization in Fabric

Getting a token proves an identity; it does not grant that identity access to
Fabric data. Configure authorization separately:

1. For service principals and managed identities, a Fabric administrator
   normally enables **Service principals can use Fabric APIs** under
   **Admin portal > Tenant settings > Developer settings**. Scope the setting
   to a dedicated Entra security group where possible. See
   [Fabric identity support](https://learn.microsoft.com/en-us/rest/api/fabric/articles/identity-support).
2. Add the user, service principal, managed identity, or an appropriate group
   to the workspace, or grant item-level permissions. Fabric
   [workspace roles](https://learn.microsoft.com/en-us/fabric/fundamentals/roles-workspaces)
   are Admin, Member, Contributor, and Viewer. Use the least-privileged role
   that permits the required operation.
3. Grant workload-specific data permissions. Workspace visibility and data
   access are not always the same permission.

The main workload considerations are:

- **Fabric discovery and item jobs.** The principal needs access to the
  workspace/item and permission for the requested read, write, or execute
  operation. Service principals and managed identities are supported only
  where the individual API says so.
- **OneLake and Delta.** Grant a workspace role, item permission, or a OneLake
  data-access role. Admin, Member, and Contributor roles can write broadly;
  granular OneLake Read or ReadWrite roles are safer for narrower access. See
  [OneLake security best practices](https://learn.microsoft.com/en-us/fabric/onelake/security/best-practices-secure-data-in-onelake).
- **SQL.** A workspace role or item Read permission permits connection, while
  SQL `GRANT`, `DENY`, and database roles control granular data access. See
  [Microsoft Entra authentication in Fabric Data Warehouse](https://learn.microsoft.com/en-us/fabric/data-warehouse/entra-id-authentication)
  and [SQL granular permissions](https://learn.microsoft.com/en-us/fabric/data-warehouse/sql-granular-permissions).
- **DAX.** Enable the **Semantic Model Execute Queries REST API** tenant
  setting. A user needs semantic-model Read and Build permissions. Service
  principals additionally need the Power BI service-principal tenant setting;
  they are not supported for models with RLS or SSO enabled. See
  [Execute Queries](https://learn.microsoft.com/en-us/rest/api/power-bi/datasets/execute-queries).
- **GraphQL.** Grant **Run Queries and Mutations** (Execute) on the GraphQL API
  and grant access to its underlying data source. A Contributor workspace role
  is convenient for development but item-level permission is narrower. See the
  [GraphQL service-principal guide](https://learn.microsoft.com/en-us/fabric/data-engineering/api-graphql-service-principal).
- **KQL.** Grant a Fabric workspace/item role or a Kusto database role such as
  `viewer`; effective access is the union of Fabric and Kusto roles. See
  [Kusto role-based access control](https://learn.microsoft.com/en-us/kusto/access-control/role-based-access-control?view=microsoft-fabric).
- **Livy.** The principal needs access to the Lakehouse and the execution/read
  permissions documented for Livy. Service-principal authentication is
  supported for app-only Spark sessions; see the
  [Livy session guide](https://learn.microsoft.com/en-us/fabric/data-engineering/get-started-api-livy-session).

When authentication succeeds but Fabric returns HTTP 403, inspect workspace,
item, and workload permissions before changing the login flow. HTTP 401 more
often indicates the wrong token resource, an expired/nonrefreshable raw token,
or a tenant/app mismatch.

## Cache and secret hygiene

`AzureAuth` stores cached user tokens in its user-specific AzureR data
directory, never in the package project. Use
`AzureAuth::list_azure_tokens()` to inspect the cache and remove only the
relevant token with `AzureAuth::delete_azure_token()`. Reserve
`AzureAuth::clean_token_directory()` for intentionally signing out of every
cached AzureR session.

Do not commit client secrets, certificates, bearer tokens, `.Renviron`, or
AzureR cache contents. In CI, inject credentials from the platform's secret
store and prefer certificate or federated credentials over client secrets
where your token provider supports them.
