---
title: "Track Eventhouse ingestion from R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Track Eventhouse ingestion from R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Kusto queued ingestion is designed for batch files that already exist in blob
storage or OneLake. fabricQueryR submits those files to an existing Eventhouse
table, retains the operation and source IDs, and turns the status response into
a failure-aware R record. The underlying REST routes are currently in preview.

## Prepare the target and source

Start with a discovered KQL database. Its record contains both the query URI
used by `fabric_kql_query()` and the ingestion URI used here:

```{r}
library(fabricQueryR)

database <- fabric_kql_databases("Telemetry workspace")[[1]]
database$ingestion_service_uri
```

The destination table must already exist. For CSV, JSON, Avro, Parquet, and ORC
workflows, create and validate a named ingestion mapping in Fabric before the R
workflow starts. General Kusto administration is intentionally outside this
API.

The source must be a Kusto storage connection string. A OneLake file can use
the workspace and item GUIDs and caller impersonation:

```{r}
source <- paste0(
  "https://onelake.dfs.fabric.microsoft.com/",
  "00000000-0000-0000-0000-000000000001/",
  "00000000-0000-0000-0000-000000000002/",
  "Files/events/2026-08-14.csv;impersonate"
)
```

The signed-in principal must be able to read that OneLake file. Other supported
storage connection strings can carry a SAS token, a managed-identity suffix, or
another authentication method documented by Kusto. Avoid writing credentialed
URLs to logs. Handles and status results redact recognized secrets, but the
submission process necessarily sends the complete source string to the trusted
Kusto ingestion endpoint.

Local paths and R data frames are not uploaded implicitly. Stage them in a
bounded format first—for example with `fabric_onelake_upload()`—and then submit
the resulting OneLake URL. This makes memory use, file format, cleanup, and
retry ownership explicit.

## Queue a tracked batch

Supply the source format and the predefined mapping. Source GUIDs are generated
when omitted and exposed on the returned handle:

```{r}
ingestion <- fabric_kql_ingest(
  database,
  table = "Events",
  sources = source,
  format = "csv",
  mapping = "EventsCsv",
  ignore_first_record = TRUE,
  tags = "source:daily-export",
  ingest_if_not_exists = "events-2026-08-14"
)

ingestion$id
ingestion$sources$source_id
ingestion$tags
```

`ingest_if_not_exists` uses stable keys without the `ingest-by:` prefix. The
function attaches the matching extent tag automatically. A later ingestion
with the same key is prevented when that tag already exists. This is the Kusto
idempotency mechanism; a source GUID primarily identifies a blob in tracking
and diagnostics.

Queued ingestion still has at-least-once delivery. In particular, concurrent
requests using the same key can race before either extent exists. Serialize
submissions sharing a key. fabricQueryR also does not retry the submission POST:
after a timeout, throttling response, or lost connection, the service might
already have accepted it, so an automatic replay could duplicate data.

The preview route accepts at most 20 blobs per request and up to 6 GB of
uncompressed data. Structured records expose known sizes without parallel
vectors:

```{r}
sources <- tibble::tibble(
  url = c(source_a, source_b),
  source_id = c(source_a_id, source_b_id),
  raw_size = c(120000000, 180000000)
)

ingestion <- fabric_kql_ingest(
  database,
  table = "Events",
  sources = sources,
  format = "parquet",
  mapping = "EventsParquet"
)
```

Keep `skip_batching = FALSE` for normal throughput. Set it only when lower
latency is more important than Kusto's normal batching efficiency. Source files
are preserved by default; `delete_after_download = TRUE` permanently removes a
successfully downloaded source and requires corresponding storage permission.

## Wait and inspect the outcome

One status snapshot is useful for a scheduler that persists operation IDs:

```{r}
snapshot <- fabric_kql_ingestion_status(ingestion)
snapshot$state
snapshot$counts
```

For an interactive or single-process batch, poll with a client-side deadline:

```{r}
result <- fabric_kql_ingestion_status(
  ingestion,
  wait = TRUE,
  timeout = 900,
  poll_interval = 2
)

result$state
result$details
```

The deadline stops only the R waiter. A `fabric_kql_ingestion_timeout`
condition includes the last observed status and explicitly leaves the service
operation running. Status GETs are safe to retry and honor service throttling
hints.

A mixed batch returns `PartiallySucceeded`; completely failed and canceled
batches have their own states. By default terminal problems raise a typed
condition containing `last_status` and the failed-source details. To handle all
states as data:

```{r}
result <- fabric_kql_ingestion_status(
  ingestion,
  wait = TRUE,
  error_on_failure = FALSE
)

failed <- subset(
  result$details,
  status %in% c("Failed", "Canceled")
)
failed[c("source_id", "error_code", "failure_status", "message")]
```

Retry a `Transient` source only after considering whether it might already have
committed. A `Permanent` failure such as a missing mapping or malformed input
requires a configuration or data fix. Query the destination after success:

```{r}
loaded <- fabric_kql_query(
  database,
  query = "Events | where ingestion_time() > ago(1h) | take 100"
)
```

The caller needs Table Ingestor permission on the destination and Database User
access. Status details for nonpublic storage also require source storage read
access. Keep the staged file until the tracked result is terminal and verified;
then remove it with the storage system's normal, explicitly authorized cleanup
workflow.
