---
title: "Working with Fabric Eventhouses (real-time data)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with Fabric Eventhouses (real-time data)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

An Eventhouse stores event, log, and time-series data in KQL databases. In this
vignette, **ingestion** means adding rows to one of its KQL tables.

Start with `fabric_kql_write_table()`: it accepts a data frame and manages the
upload, ingestion status, and cleanup for you. Later sections cover larger
Arrow inputs and the lower-level queued-ingestion workflow for files that are
already in OneLake or blob storage. The underlying ingestion REST routes are
currently in preview.

## Find a KQL database

Start with a discovered KQL database. Its record contains both the query URI
used by `fabric_kql_query()` and the ingestion URI used here:

```{r, eval = FALSE}
library(fabricQueryR)

database <- fabric_kql_databases("Telemetry workspace")[[1]]
database$ingestion_service_uri
```

## Write a small data frame in one call

Supply a destination table and an ordinary data frame or tibble. Use a new test
table while learning:

```{r, eval = FALSE}
written <- fabric_kql_write_table(
  database,
  table = "Events",
  data = data.frame(
    id = 1:3,
    category = c("A", "B", "A"),
    amount = c(10.5, 20, 30.5)
  ),
  create_if_missing = TRUE,
  ingest_if_not_exists = "r-events-2026-08-14"
)

written$status$state
written$rows
written$staging_retained
```

The function writes temporary Parquet data, uploads it to Fabric, queues the
ingestion, waits for a final status, and deletes staging only after confirmed
success. `create_if_missing = TRUE` creates a basic table from the R object's
columns when needed.

With `create_if_missing = TRUE`, the writer sends Kusto's idempotent `.create
table` command before staging. A missing target is inferred from the Arrow
schema; an existing target is returned unchanged. Common logical, integer,
double, character, date/time, decimal, and nested Arrow types are mapped to
Kusto scalar types. Use a named `column_types` vector with one value per source
field when those inferred types are not appropriate.

Parquet identity mapping is case-sensitive. Source field names and types must
match an existing KQL table; otherwise supply a predefined Parquet `mapping`.
Factors are written as strings. Convert complex and `difftime` columns
explicitly.

Arrow Tables and RecordBatches work without an R data-frame conversion. For
data larger than memory, pass a lazy Dataset, Scanner, Arrow dplyr query,
RecordBatchReader, or an Arrow-compatible stream returned by a package query:

```{r, eval = FALSE}
dataset <- arrow::open_dataset("local-parquet-directory")

written <- fabric_kql_write_table(
  database,
  table = "Events",
  data = dataset,
  mapping = "EventsParquet",
  create_if_missing = FALSE
)
```

The dataset is scanned one record batch at a time into bounded temporary
Parquet parts and then uploaded in bounded chunks. It is never collected into a
data frame or Arrow Table. A supplied RecordBatchReader is single-use and is
exhausted by the write. Use `target_file_size` for a soft byte boundary or
`max_rows_per_file` for an exact row boundary. The serialized batch must remain
within the ingestion service's advertised size and blob-count limits.

The function uses the OneLake folder returned by
`/v1/rest/ingestion/configuration`. If an administrator provides a different
writable location, pass its full trusted `Files/` URI as `staging_folder`.
Ambiguous submission and polling failures always retain staging. Confirmed
terminal failures retain it by default; set
`keep_staging_on_failure = FALSE` only when automatic removal is appropriate.
An upload transport error can make the final atomic rename ambiguous; its
condition reports the unique path with `staging_retained = NA` for inspection.

## Queue a tracked batch

Use the lower-level `fabric_kql_ingest()` when the source file already exists
in OneLake or supported blob storage. It does not upload a local file or
serialize an R object, and the destination table must already exist.

For CSV, JSON, Avro, Parquet, and ORC workflows, create and validate a named
ingestion mapping in Fabric before submission. A OneLake source can use the
workspace and item GUIDs with caller impersonation:

```{r, eval = FALSE}
source <- paste0(
  "https://onelake.dfs.fabric.microsoft.com/",
  "00000000-0000-0000-0000-000000000001/",
  "00000000-0000-0000-0000-000000000002/",
  "Files/events/2026-08-14.csv;impersonate"
)
```

The signed-in identity must be able to read the file. Other Kusto storage
connection strings can contain a SAS token, managed-identity suffix, or another
documented authentication method. Do not print or log a credential-bearing
source string. The complete value must be sent to the trusted ingestion
endpoint even though returned objects redact recognized secrets.

Supply the source format and the predefined mapping. Source GUIDs are generated
when omitted and exposed on the returned handle:

```{r, eval = FALSE}
ingestion <- fabric_kql_ingest(
  database,
  table = "Events",
  sources = source,
  format = "csv",
  mapping = "EventsCsv",
  ignore_first_record = TRUE,
  tags = "source:daily-export",
  ingest_if_not_exists = "events-2026-08-14"
)

ingestion$id
ingestion$sources$source_id
ingestion$tags
```

`ingest_if_not_exists` uses stable keys without the `ingest-by:` prefix. The
function attaches the matching extent tag automatically. A later ingestion
with the same key is prevented when that tag already exists. This is the Kusto
idempotency mechanism; a source GUID primarily identifies a blob in tracking
and diagnostics.

Queued ingestion still has at-least-once delivery. In particular, concurrent
requests using the same key can race before either extent exists. Serialize
submissions sharing a key. fabricQueryR also does not retry the submission POST:
after a timeout, throttling response, or lost connection, the service might
already have accepted it, so an automatic replay could duplicate data.

The preview route accepts at most 20 blobs per request and up to 6 GB of
uncompressed data. Structured records expose known sizes without parallel
vectors:

```{r, eval = FALSE}
sources <- tibble::tibble(
  url = c(source_a, source_b),
  source_id = c(source_a_id, source_b_id),
  raw_size = c(120000000, 180000000)
)

ingestion <- fabric_kql_ingest(
  database,
  table = "Events",
  sources = sources,
  format = "parquet",
  mapping = "EventsParquet"
)
```

Keep `skip_batching = FALSE` for normal throughput. Set it only when lower
latency is more important than Kusto's normal batching efficiency. Source files
are preserved by default; `delete_after_download = TRUE` permanently removes a
successfully downloaded source and requires corresponding storage permission.

## Wait and inspect the outcome

One status snapshot is useful for a scheduler that persists operation IDs:

```{r, eval = FALSE}
snapshot <- fabric_kql_ingestion_status(ingestion)
snapshot$state
snapshot$counts
```

For an interactive or single-process batch, poll with a client-side deadline:

```{r, eval = FALSE}
result <- fabric_kql_ingestion_status(
  ingestion,
  wait = TRUE,
  timeout = 900,
  poll_interval = 2
)

result$state
result$details
```

The deadline stops only the R waiter. A `fabric_kql_ingestion_timeout`
condition includes the last observed status and explicitly leaves the service
operation running. Status GETs are safe to retry and honor service throttling
hints.

A mixed batch returns `PartiallySucceeded`; completely failed and canceled
batches have their own states. By default terminal problems raise a typed
condition containing `last_status` and the failed-source details. To handle all
states as data:

```{r, eval = FALSE}
result <- fabric_kql_ingestion_status(
  ingestion,
  wait = TRUE,
  error_on_failure = FALSE
)

failed <- subset(
  result$details,
  status %in% c("Failed", "Canceled")
)
failed[c("source_id", "error_code", "failure_status", "message")]
```

Retry a `Transient` source only after considering whether it might already have
committed. A `Permanent` failure such as a missing mapping or malformed input
requires a configuration or data fix. Query the destination after success:

```{r, eval = FALSE}
loaded <- fabric_kql_query(
  database,
  query = "Events | where ingestion_time() > ago(1h) | take 100"
)
```

The caller needs Table Ingestor permission on the destination and Database User
access. Status details for nonpublic storage also require source storage read
access. For manual `fabric_kql_ingest()` workflows, keep the staged file until
the tracked result is terminal and verified. `fabric_kql_write_table()` applies
that retention rule automatically and reports the full retained path whenever
manual recovery is required.

## Export a large KQL result to OneLake

`fabric_kql_query()` is the right interface when the result belongs in R. When
the result is too large for the client-result channel or should remain in
Fabric, `fabric_kql_export()` runs Kusto's service-side export and writes the
first result set directly to storage:

```{r, eval = FALSE}
lakehouse <- fabric_lakehouses("Telemetry workspace")[[1]]

exported <- fabric_kql_export(
  database,
  query = "Events | where observed_at > ago(7d)",
  destination = lakehouse,
  path = "Files/exports/events-weekly",
  format = "parquet",
  name_prefix = "events",
  compression_type = "snappy"
)

exported$state
exported$records
exported$artifacts
```

A discovered item uses a OneLake GUID path with `;impersonate`; the principal
therefore needs write access to the destination. A complete documented Kusto
storage connection string can be supplied instead. Credential-bearing strings
are obfuscated in Kusto command telemetry and redacted from returned R objects.

The function always uses `async`, polls `.show operations`, and calls
`.show operation ... details` only after the operation reports `Completed`.
The submission itself is never replayed. Kusto documents that ordinary exports
do not retry and that artifacts already written by a failed export remain and
must be treated as incomplete. Timeout and failure conditions carry the
operation ID and safe destination so the operation can be inspected before a
new export is submitted.
