---
title: "Get started with fabricQueryR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with fabricQueryR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Microsoft Fabric is a collection of services for storing, transforming, and
reporting on data. `fabricQueryR` lets you use many of those services from R.
You can bring Fabric data into an R data frame, send R data to Fabric, or
use R to start work that runs inside Fabric.

This guide introduces the few of the Fabric concepts you need for a first successful
workflow. 

## The Fabric objects you will see

A *workspace* is a shared area that contains Fabric items. An *item* is a
resource inside a workspace, such as a Lakehouse, Warehouse, semantic model, 
or notebook.

The most common data items have different purposes:

| Item | Think of it as | A common R task |
|---|---|---|
| Lakehouse | Files plus managed data tables | Read or write a table or file |
| Warehouse | A relational SQL database | Query or load business tables |
| Eventhouse | A database for event and time-series data | Query with KQL or ingest events |
| Semantic model | Report-ready tables, relationships, and calculations | Query with DAX or refresh the model |
| API for GraphQL | A structured API in front of Fabric data | Request selected fields |

*OneLake* is the storage layer shared by Fabric items. In a Lakehouse, the
`Files/` area contains ordinary files and the `Tables/` area contains managed
Delta tables. Delta is a storage format that supports efficient reads and writes,
schema evolution, and transactional consistency.

## Sign in

Install and load the package, 
then set your organization's Microsoft Entra tenant ID:

```{r, eval = FALSE}
install.packages("fabricQueryR")

library(fabricQueryR)
Sys.setenv(FABRICQUERYR_TENANT_ID = "<your-tenant-id>")

# Optional, if your organization requires an approved application:
Sys.setenv(FABRICQUERYR_CLIENT_ID = "<your-app-client-id>")
```

The first Fabric call may open a browser. Sign in with the same work or school
account that you use in the Fabric portal. 

If your organization requires an approved application, 
your administrator may also give you a client ID to set as `FABRICQUERYR_CLIENT_ID`.

The [authentication vignette](authentication.html) explains this setup and
the different ways to authenticate in more detail.

## Find a workspace and an item

Start by listing the workspaces that your account can access:

```{r, eval = FALSE}
# List all workspaces you can access
workspaces <- fabric_workspaces()
```

The result is a list of workspaces you have access to. 
If the list is empty, check that your account has been granted access to a workspace in the Fabric portal.
If the list is not empty, select a specific workspace:

```{r, eval = FALSE}
# Select the first workspace in the list
workspace <- workspaces[[1L]]
workspace$displayName
```

For a script that will run repeatedly,
selecting by exact name is more robust:

```{r, eval = FALSE}
# Select a workspace by name
workspace <- fabric_workspaces()[["Analytics workspace"]]
```

Now list all items in the workspace, or ask directly for items of a specific type:

```{r, eval = FALSE}
# List all items in the workspace
items <- fabric_items(workspace)
items

# List only Lakehouses in the workspace
lakehouses <- fabric_lakehouses(workspace)
lakehouse <- lakehouses[[1L]]
lakehouse$displayName
```

A discovered item contains metadata such as its name and type.
You can pass the discovered item to other functions of this package (see below).

## Complete a first read

If the workspace contains a Lakehouse, 
reading one managed table is a simple first workflow:

```{r, eval = FALSE}
# List the tables in the Lakehouse
tables <- fabric_lakehouse_tables(lakehouse)
tables[c("schema", "name", "type")]

# Select the first table and read a small number of rows
first_table <- tables[1L, ]
rows <- fabric_lakehouse_read_table(
  lakehouse,
  first_table,
  limit = 100L
)

# Show the first few rows
head(rows)
```

The result is a tibble, which can be used with base R, `dplyr`, plotting
packages, or other familiar R tools. `limit = 100L` keeps this first request
small while you confirm that access and table selection are correct.

## Choose the next guide

There are often several valid ways to move the same data. 
The vignettes below compare the options and show how to use them.
Continue with one of the following vignettes:

- [Bring Fabric data into R](reading-data.html) compares SQL, Lakehouse,
  Warehouse, Eventhouse, semantic-model, OneLake-file, GraphQL, and Spark
  reads
- [Bring R data into Microsoft Fabric](ingesting-data.html) compares ways to
  send an R object or an existing file to Fabric
- [Working with Fabric LakeHouses and OneLake](onelake-and-lakehouse.html)
  explains Lakehouse files and tables in more detail
- [Working with Livy (Spark)](spark-with-livy.html) introduces remote Spark work
  after the simpler read and write paths
