---
title: "Working with GraphQL"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with GraphQL}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

A Fabric API for GraphQL is a structured doorway to data selected by the API's
owner. A GraphQL query names the fields you want, and the response follows the
same nested shape. This can be convenient when you need an approved API view
instead of direct access to the underlying Lakehouse, Warehouse, or SQL
Database.

This guide starts with one small query. It then introduces schema inspection,
pagination, and collection into a tibble.

Start with a discovered API so its endpoint and workspace travel together:

```{r, eval = FALSE}
library(fabricQueryR)

api <- fabric_graphql_apis("Analytics workspace")[[1]]
```

The caller needs *Run Queries and Mutations* permission on the API. With SSO
connectivity, the caller also needs access to the underlying data source. The
package uses the delegated `GraphQLApi.Execute.All` scope for interactive users
and the Fabric audience for service-principal authentication.

## Run a first query

A GraphQL document is one character string. The field names depend on how the
Fabric API was configured, so replace `products`, `id`, and `name` with fields
from your API:

```{r, eval = FALSE}
response <- fabric_graphql_query(
  api,
  query = "{ products { items { id name } } }"
)

response$data$products$items
response$errors
```

GraphQL can return useful data and service errors together. 'fabricQueryR' keeps
them separate so you can inspect both. When this small request works, use the
next sections to discover fields and retrieve more than one page.

## Inspect the available fields

Microsoft Fabric disables runtime introspection by default. Only a workspace
admin can enable it under *API Settings > Introspection*. Once enabled, the
standard introspection response is available as a nested R list:

```{r, eval = FALSE}
schema <- fabric_graphql_schema(api)

schema$queryType$name
vapply(schema$types, `[[`, character(1), "name")
```

`fabric_graphql_schema()` stops with a
`fabric_graphql_introspection_error` when the service does not return a complete
schema. Its message points to the administrator setting and the portal's
*Export schema* alternative. Export remains available when runtime
introspection must stay disabled.

## Read more than one page

Fabric normally represents a generated collection with `items`,
`hasNextPage`, and `endCursor`. Request all three pieces needed by the workflow
and use a stable explicit ordering when pages must be repeatable:

```{r, eval = FALSE}
pages <- fabric_graphql_paginate(
  api,
  query = paste(
    "query Products($first: Int!, $after: String) {",
    "  products(first: $first, after: $after, orderBy: {id: ASC}) {",
    "    items {",
    "      id",
    "      name",
    "      category { id name }",
    "      tags",
    "    }",
    "    hasNextPage",
    "    endCursor",
    "  }",
    "}"
  ),
  variables = list(first = 100L, after = NULL),
  operation_name = "Products",
  next_cursor = fabric_graphql_cursor("products"),
  idempotent = TRUE
)
```

The cursor is opaque. `fabric_graphql_cursor()` only reads it from the
configured connection path and feeds it back through the `after` variable; it
does not interpret the value.

## Build an analysis-ready tibble

Select the row array explicitly relative to each page's `data` field:

```{r, eval = FALSE}
products <- fabric_graphql_collect(pages, c("products", "items"))

products
attr(products, "complete")
attr(products, "page_count")
attr(products, "errors")
```

Scalar fields become ordinary columns. A field that first appears on a later
page is added to the union of columns, with typed `NA` values in earlier rows.
Nested objects and arrays remain list-columns rather than being flattened:

```{r, eval = FALSE}
products$category[[1]]
products$tags[[1]]
```

JSON whole numbers beyond R's exact double range are decoded as character
values. The collector preserves those strings. If the same field has smaller R
integer values on another page, it promotes those smaller values to character
too; it never converts the large value to an inexact double. Other incompatible
scalar changes stop with an error so a schema change cannot silently alter the
analysis.

GraphQL can return data and errors together. The rows remain usable, while the
combined service errors stay in `attr(products, "errors")` and their count is
shown when the tibble prints.

## Treat incomplete pagination as partial data

`fabric_graphql_paginate()` marks completion only after the cursor callback
reports that no next page exists. Reaching `max_pages` raises a
`fabric_graphql_pagination_error` whose `pages` field is marked incomplete. If
those pages are passed to `fabric_graphql_collect()`, collection raises a
`fabric_graphql_collection_error` instead of returning an apparently complete
tibble:

```{r, eval = FALSE}
tryCatch(
  fabric_graphql_collect(incomplete_pages, c("products", "items")),
  fabric_graphql_collection_error = function(error) {
    partial <- error$partial_data
    attr(partial, "complete")
    partial
  }
)
```

Fabric currently defaults each page to 100 items and limits pagination to
100,000 items. A response is limited to 64 MB, one request to 100 seconds, and
query nesting to 10 levels. Use smaller pages and stable filtered partitions
for large results rather than treating a service-limit response as completion.

See Microsoft's documentation for
[introspection and schema export](https://learn.microsoft.com/en-us/fabric/data-engineering/api-graphql-introspection-schema-export),
[GraphQL limits](https://learn.microsoft.com/en-us/fabric/data-engineering/api-graphql-limits),
and the generated
[pagination shape](https://learn.microsoft.com/en-us/fabric/data-engineering/api-graphql-aggregations).
