---
title: "Bring R data into Microsoft Fabric"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bring R data into Microsoft Fabric}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

'fabricQueryR' can also be used to ingest data into Microsoft Fabric.
In this guide, it is shown how you can move R data frames, Arrow objects, or files
into Microsoft Fabric.

## Choose a destination

| What you want in Fabric | Start with | Good fit |
|---|---|---|
| A managed Lakehouse Delta table | `fabric_lakehouse_write_table()` | General analytics and data-engineering tables |
| An ordinary file in OneLake | `fabric_onelake_write_file()` | Exchange files, exports, and non-tabular artifacts |
| A relational Warehouse table | `fabric_warehouse_write_table()` | SQL reporting and warehouse workloads |
| An Eventhouse KQL table | `fabric_kql_write_table()` | Event, log, and time-series data |
| A Lakehouse table from files already in `Files/` | `fabric_lakehouse_load_table()` | Existing CSV or Parquet staging data |

For a first ingestion, a Lakehouse table is the most direct general-purpose
workflow. The high-level writers accept ordinary data frames and handle their
own temporary Parquet staging.

## Prepare a small R dataframe

For demonstration purposes, 
we'll first create a small R dataframe and select a Lakehouse in a workspace:

```{r, eval = FALSE}
library(fabricQueryR)

orders <- data.frame(
  order_id = 1:3,
  order_date = as.Date(c("2026-08-12", "2026-08-13", "2026-08-14")),
  amount = c(10.50, 20, 30.25)
)

workspaces <- fabric_workspaces()
matches <- Filter(
  \(x) identical(x$displayName, "Analytics workspace"),
  workspaces
)
stopifnot(length(matches) == 1L)
workspace <- matches[[1L]]

lakehouse <- fabric_lakehouses(workspace)[[1L]]
```

## Write a Lakehouse table

```{r, eval = FALSE}
write_result <- fabric_lakehouse_write_table(
  lakehouse,
  table = "orders_from_r",
  data = orders,
  mode = "Overwrite"
)

write_result$rows
write_result$staging_retained
```

The function writes temporary Parquet files, loads them as a managed Delta
table, waits for Fabric to finish, and removes successful staging files. It can
create the destination table; Fabric infers its columns from the source.

Read back a few rows to verify the result:

```{r, eval = FALSE}
check <- fabric_lakehouse_read_table(
  lakehouse,
  table = "orders_from_r",
  limit = 10L
)

check
```

Use `mode = "Append"` only when the source columns are compatible with an
existing table. `mode = "Overwrite"` replaces the table through Fabric's
managed load behavior.

## Write an ordinary OneLake file

A file is different from a managed table. Choose this route when another
process expects a specific file or when the content is not tabular:

```{r, eval = FALSE}
fabric_onelake_write_file(
  workspace,
  lakehouse,
  path = "Files/exports/orders.parquet",
  data = orders
)
```

`fabric_onelake_write_file()` serializes supported R or Arrow objects. Use
`fabric_onelake_upload()` when a file already exists on local disk:

```{r, eval = FALSE}
fabric_onelake_upload(
  workspace,
  lakehouse,
  path = "Files/incoming/orders.csv",
  source = "orders.csv"
)
```

Do not upload directly below a managed table's `Tables/` directory. Delta
tables contain a transaction log and must be changed through a table-aware
writer.

## Write a Warehouse table

A Warehouse writer uses a Lakehouse `Files/` directory for temporary staging,
then asks the Warehouse to load it efficiently:

```{r, eval = FALSE}
warehouse <- fabric_warehouses(workspace)[[1L]]

warehouse_result <- fabric_warehouse_write_table(
  warehouse,
  table = "orders_from_r",
  data = orders,
  staging_lakehouse = lakehouse,
  schema = "dbo",
  create_if_missing = TRUE,
  mode = "Append"
)
```

For a missing table, Fabric can infer a basic definition. Pre-create the table
when exact SQL types, lengths, constraints, or grants matter.

[Working with Fabric Warehouses](warehouse.html) explains overwrite
choices, transactions, and larger Arrow inputs.

## Write an Eventhouse table

Use Eventhouse for event or time-series data that will be queried with KQL:

```{r, eval = FALSE}
kql_database <- fabric_kql_databases(workspace)[[1L]]

kql_result <- fabric_kql_write_table(
  kql_database,
  table = "OrdersFromR",
  data = orders,
  create_if_missing = TRUE
)

kql_result$status$state
```

The high-level writer stages the R object, submits tracked ingestion, waits,
and cleans up after confirmed success.

[Working with Fabric Eventhouses (real-time data)](eventhouse-ingestion.html)
covers predefined mappings, existing storage files, idempotency keys, and
failure recovery.

## Load a file that is already in a Lakehouse

If CSV or Parquet data already exists below the same Lakehouse's `Files/` area,
you can load it without downloading it to R:

```{r, eval = FALSE}
operation <- fabric_lakehouse_load_table(
  lakehouse,
  table = "orders_from_file",
  path = "Files/incoming/orders.csv",
  format = "Csv",
  header = TRUE,
  mode = "Overwrite"
)

completed <- fabric_operation_wait(operation, timeout = 900)
```

This route is useful for file-based pipelines. It does not upload a local file;
use `fabric_onelake_upload()` first when necessary.

## Scale up with Arrow

When you are moving larger amounts of data, consider using Arrow.

The Lakehouse, Warehouse, and Eventhouse writers accept Arrow Tables and
RecordBatches. They can also consume lazy Arrow Datasets, Scanners, queries,
and streams in batches, without collecting the full input as an R data frame:

```{r, eval = FALSE}
dataset <- arrow::open_dataset("local-parquet-directory")

fabric_lakehouse_write_table(
  lakehouse,
  table = "large_orders",
  data = dataset
)
```
