---
title: "Bring R data into Microsoft Fabric"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bring R data into Microsoft Fabric}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

'fabricQueryR' helps you send data from R to Lakehouses, Warehouses,
Eventhouses, and OneLake. The source can be an ordinary R data frame, an Arrow
object, or a file that already exists on disk or in Fabric.

The right writer depends on how the data will be used after it arrives. This
guide starts with a small data frame and a managed Lakehouse table, then shows
the other common destinations. Larger Arrow workflows come last.

## Choose a destination

| What you want in Fabric | Start with | Good fit |
|---|---|---|
| A managed Lakehouse Delta table | `lakehouse$write_table()` (`fabric_lakehouse_write_table()`) | General analytics and data-engineering tables |
| An ordinary file in OneLake | `lakehouse$onelake_write_file()` (`fabric_onelake_write_file()`) | Exchange files, exports, and non-tabular artifacts |
| A relational Warehouse table | `warehouse$write_table()` (`fabric_warehouse_write_table()`) | SQL reporting and warehouse workloads |
| An Eventhouse KQL table | `kql_database$write_table()` (`fabric_kql_write_table()`) | Event, log, and time-series data |
| A Lakehouse table from files already in `Files/` | `lakehouse$load_table()` (`fabric_lakehouse_load_table()`) | Existing CSV or Parquet staging data |

For a first ingestion, a Lakehouse table is the most direct general-purpose
workflow. The high-level writers accept ordinary data frames and handle their
own temporary Parquet staging.

## Prepare a small R data frame

Create a small data frame and select a Lakehouse in a workspace:

```{r, eval = FALSE}
library(fabricQueryR)

orders <- data.frame(
  order_id = 1:3,
  order_date = as.Date(c("2026-08-12", "2026-08-13", "2026-08-14")),
  amount = c(10.50, 20, 30.25)
)

workspaces <- fabric_workspaces()
matches <- Filter(
  \(x) identical(x$displayName, "Analytics workspace"),
  workspaces
)
stopifnot(length(matches) == 1L)
workspace <- matches[[1L]]

lakehouse <- workspace$lakehouses()[[1L]]
```

`workspace` and `lakehouse` are read-only R6 objects returned by discovery.
Read their Fabric fields through `$`; their methods use the IDs and credential
needed for the next operation. `$lakehouses()` is the workspace method for
`fabric_lakehouses()`.

## Write a Lakehouse table

Call `$write_table()` (`fabric_lakehouse_write_table()`) on the discovered
Lakehouse:

```{r, eval = FALSE}
write_result <- lakehouse$write_table(
  table = "orders_from_r",
  data = orders,
  mode = "Overwrite"
)

write_result$rows
write_result$staging_retained
```

The function writes temporary Parquet files, loads them as a managed Delta
table, waits for Fabric to finish, and removes successful staging files. It can
create the destination table; Fabric infers its columns from the source.

Read back a few rows with `$read_table()`
(`fabric_lakehouse_read_table()`) to verify the result:

```{r, eval = FALSE}
check <- lakehouse$read_table(
  table = "orders_from_r",
  limit = 10L
)

check
```

Use `mode = "Append"` only when the source columns are compatible with an
existing table. `mode = "Overwrite"` replaces the table through Fabric's
managed load behavior.

## Write an ordinary OneLake file

A file is different from a managed table. Choose this route when another
process expects a specific file or when the content is not tabular:

```{r, eval = FALSE}
lakehouse$onelake_write_file(
  path = "Files/exports/orders.parquet",
  data = orders
)
```

`$onelake_write_file()` (`fabric_onelake_write_file()`) serializes supported R
or Arrow objects. Use `$onelake_upload()` (`fabric_onelake_upload()`) when a
file already exists on local disk:

```{r, eval = FALSE}
lakehouse$onelake_upload(
  path = "Files/incoming/orders.csv",
  source = "orders.csv"
)
```

Do not upload directly below a managed table's `Tables/` directory. Delta
tables contain a transaction log and must be changed through a table-aware
writer.

## Write a Warehouse table

A Warehouse writer uses a Lakehouse `Files/` directory for temporary staging,
then asks the Warehouse to load it efficiently. Discover Warehouses with
`$warehouses()` (`fabric_warehouses()`), then write with `$write_table()`
(`fabric_warehouse_write_table()`):

```{r, eval = FALSE}
warehouse <- workspace$warehouses()[[1L]]

warehouse_result <- warehouse$write_table(
  table = "orders_from_r",
  data = orders,
  staging_lakehouse = lakehouse,
  schema = "dbo",
  create_if_missing = TRUE,
  mode = "Append"
)
```

For a missing table, Fabric can infer a basic definition. Pre-create the table
when exact SQL types, lengths, constraints, or grants matter.

[Working with Fabric Warehouses](warehouse.html) explains overwrite
choices, transactions, and larger Arrow inputs.

## Write an Eventhouse table

Use Eventhouse for event or time-series data that will be queried with KQL.
Discover KQL databases with `$kql_databases()` (`fabric_kql_databases()`), then
write with `$write_table()` (`fabric_kql_write_table()`):

```{r, eval = FALSE}
kql_database <- workspace$kql_databases()[[1L]]

kql_result <- kql_database$write_table(
  table = "OrdersFromR",
  data = orders,
  create_if_missing = TRUE
)

kql_result$status$state
```

The high-level writer stages the R object, submits tracked ingestion, waits,
and cleans up after confirmed success.

[Working with Fabric Eventhouses (real-time data)](eventhouse-ingestion.html)
covers predefined mappings, existing storage files, idempotency keys, and
failure recovery.

## Load a file that is already in a Lakehouse

If CSV or Parquet data already exists below the same Lakehouse's `Files/` area,
you can load it without downloading it to R. The `$load_table()` method calls
`fabric_lakehouse_load_table()`:

```{r, eval = FALSE}
operation <- lakehouse$load_table(
  table = "orders_from_file",
  path = "Files/incoming/orders.csv",
  format = "Csv",
  header = TRUE,
  mode = "Overwrite"
)

completed <- fabric_operation_wait(operation, timeout = 900)
```

This route is useful for file-based pipelines. It does not upload a local file;
use `$onelake_upload()` (`fabric_onelake_upload()`) first when necessary.

## Scale up with Arrow

When you are moving larger amounts of data, consider using Arrow.

The Lakehouse, Warehouse, and Eventhouse writers accept Arrow Tables and
RecordBatches. They can also consume lazy Arrow Datasets, Scanners, queries,
and streams in batches, without collecting the full input as an R data frame:

This example again uses `$write_table()`
(`fabric_lakehouse_write_table()`):

```{r, eval = FALSE}
dataset <- arrow::open_dataset("local-parquet-directory")

lakehouse$write_table(
  table = "large_orders",
  data = dataset
)
```
