---
title: "Automate and inspect Microsoft Fabric jobs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Automate and inspect Microsoft Fabric jobs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

A Fabric **job** is a run of an item such as a notebook, data pipeline, or Spark
job definition. fabricQueryR can start a run, wait for it, inspect recent runs,
and manage recurring schedules.

Start with one on-demand run. Add a schedule only after that run succeeds and
its parameters and runtime are understood. This guide uses a notebook, but the
same pattern applies to other supported item types.

```{r, eval = FALSE}
library(fabricQueryR)

notebook <- fabric_notebooks("Analytics workspace")[[1]]
```

## Run once and inspect history

Start an on-demand job and wait for its terminal state:

```{r, eval = FALSE}
job <- fabric_job_run(
  notebook,
  parameters = list(run_date = Sys.Date(), full_load = FALSE)
)
result <- fabric_job_wait(job, timeout = 900, cancel_on_timeout = TRUE)
result$status
```

The first call returns immediately with a job handle. The wait call checks
Fabric until the job finishes or the 15-minute local deadline is reached.
`cancel_on_timeout = TRUE` asks Fabric to cancel the run if that deadline is
exceeded.

Running jobs needs item execute permission. Reading history and schedules
needs item read permission. Creating or updating a schedule needs item execute
and read-write permissions; deletion needs item read-write permission. A
workspace Contributor role or direct item Write permission is normally needed
to change schedules.

`fabric_job_instances()` follows all Fabric continuation pages. Most item types
retain at most 100 recently completed runs, in addition to active runs.

```{r, eval = FALSE}
history <- fabric_job_instances(notebook)

history[[1]]$invoke_type
history[[1]]$status
history[[1]]$start_time
history[[1]]$failure_reason
```

History entries use the same `fabric_job_instance` contract as on-demand status
results. They can be refreshed directly without copying an instance ID:

```{r, eval = FALSE}
latest <- fabric_job_status(history[[1]])
```

Fabric can add job statuses and invocation types over time. Unknown values stay
unchanged in the normalized fields and the complete response is available in
`raw`.

## Build schedule configurations

Fabric currently documents four REST configuration types. Schedule boundaries
are absolute UTC instants. Daily, weekly, and monthly clock times are local to a
Windows time-zone identifier, which lets Fabric apply daylight-saving rules.
They never depend on the computer running R.

Cron is Fabric's name for a minute interval, not a cron expression:

```{r, eval = FALSE}
every_30_minutes <- fabric_job_schedule_config(
  "Cron",
  start_time = "2026-10-01T00:00:00Z",
  end_time = "2027-10-01T00:00:00Z",
  time_zone = "UTC",
  interval = 30L
)
```

Daily and weekly schedules accept one or more `HH:MM` values. Use a Windows
identifier such as `W. Europe Standard Time`, not an IANA identifier such as
`Europe/Amsterdam`.

```{r, eval = FALSE}
daily <- fabric_job_schedule_config(
  "Daily",
  start_time = "2026-10-01T00:00:00Z",
  end_time = "2027-10-01T00:00:00Z",
  time_zone = "W. Europe Standard Time",
  times = c("08:30", "17:00")
)

weekly <- fabric_job_schedule_config(
  "Weekly",
  start_time = "2026-10-01T00:00:00Z",
  end_time = "2027-10-01T00:00:00Z",
  time_zone = "W. Europe Standard Time",
  times = "07:30",
  weekdays = c("Monday", "Thursday")
)
```

Monthly schedules repeat every 1 through 12 months. Select either a numbered
day or an ordinal weekday. Fabric skips months where a numbered day is invalid.

```{r, eval = FALSE}
month_end_review <- fabric_job_schedule_config(
  "Monthly",
  start_time = "2026-10-01T00:00:00Z",
  end_time = "2027-10-01T00:00:00Z",
  time_zone = "W. Europe Standard Time",
  times = "09:00",
  recurrence = 1L,
  week_index = "Fourth",
  weekday = "Friday"
)
```

## Create, list, update, and disable

An item can currently have at most 20 schedules. Create one from a validated
configuration, then list all of the schedules for that job type:

```{r, eval = FALSE}
schedule <- fabric_job_schedule_create(notebook, weekly, enabled = TRUE)
schedules <- fabric_job_schedules(notebook)
```

The Fabric PATCH endpoint requires both `enabled` and the complete schedule
configuration. For a partial R update, fabricQueryR first reads the current
record and preserves fields that were not supplied:

```{r, eval = FALSE}
disabled <- fabric_job_schedule_update(
  notebook,
  schedule,
  enabled = FALSE
)

restarted <- fabric_job_schedule_update(
  notebook,
  schedule,
  enabled = TRUE
)
```

Fabric may automatically disable a schedule after repeated failures. Updating
it with `enabled = TRUE` restarts it. The published REST response exposes the
`enabled` flag but does not yet guarantee a separate auto-disable reason, so
`auto_disabled` is `NA` unless the service returns an explicit marker. Inspect
`raw` for any newer workload-specific fields.

Some job types accept static `executionData`. Its schema belongs to that
workload, so provide the exact documented object:

```{r, eval = FALSE}
special <- fabric_job_schedule_create(
  workload_item,
  daily,
  job_type = "DefaultJob",
  execution_data = list(
    tableName = "Sales",
    optimizeSettings = list(vOrder = TRUE)
  )
)
```

Omitting `execution_data` during update, or leaving it as `NULL`, preserves the
current value. Supply a named list to replace it.

## Future schedule types and cleanup

For a schedule type added after this package version, pass the documented JSON
shape as a named list. Known types are validated; unknown types are sent without
discarding their fields and returned in `configuration` and `raw`.

```{r, eval = FALSE}
future_configuration <- list(
  type = "FutureScheduleType",
  futureProperty = list(value = "documented by Fabric")
)
```

Deletion is permanent and never inferred from a create or update conflict. It
requires explicit confirmation:

```{r, eval = FALSE}
fabric_job_schedule_delete(notebook, schedule, confirm = TRUE)
```

Microsoft documents scheduler throttling, maximum job duration and concurrency,
schedule expiry after prolonged user inactivity, and auto-disable behavior in
the [Fabric job scheduler
guide](https://learn.microsoft.com/en-us/fabric/fundamentals/job-scheduler).
