---
title: "Automate and inspect Microsoft Fabric jobs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Automate and inspect Microsoft Fabric jobs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

A Fabric *job* is a run of an item such as a notebook, data pipeline, or Spark
job definition. 'fabricQueryR' can start a run, wait for it, inspect recent runs,
and manage recurring schedules.

Start with one on-demand run. Add a schedule only after that run succeeds and
its parameters and runtime are understood. This guide uses a notebook, but the
same pattern applies to other supported item types.

```{r, eval = FALSE}
library(fabricQueryR)

notebook <- fabric_notebooks("Analytics workspace")[[1]]
```

## Run once and inspect history

Start an on-demand job and wait for its terminal state:

```{r, eval = FALSE}
job <- fabric_job_run(
  notebook,
  parameters = list(run_date = Sys.Date(), full_load = FALSE)
)
result <- fabric_job_wait(job, timeout = 900, cancel_on_timeout = TRUE)
result$status
```

The first call returns immediately with a job handle. The wait call checks
Fabric until the job finishes or the 15-minute local deadline is reached.
`cancel_on_timeout = TRUE` asks Fabric to cancel the run if that deadline is
exceeded.

Notebook submission uses Fabric's Core Job Scheduler route so run parameters
and compute settings are applied without relying on the workload submission
route scheduled for deprecation. If a Notebook workflow needs the beta status
fields, such as its exit value, opt in explicitly:

```{r, eval = FALSE}
detailed <- fabric_job_status(
  job,
  notebook_details = TRUE,
  respect_retry_after = FALSE
)
detailed$exit_value
```

Running a job needs Execute permission. Reading history needs Read permission,
while changing schedules normally needs Write access. If an on-demand run works
but schedule creation does not, ask the item owner to check your Write access.

```{r, eval = FALSE}
history <- fabric_job_instances(notebook)

history[[1]]$invoke_type
history[[1]]$status
history[[1]]$start_time
history[[1]]$failure_reason
```

Refresh a history entry directly without copying its instance ID:

```{r, eval = FALSE}
latest <- fabric_job_status(history[[1]])
```

## Build schedule configurations

Build and validate a schedule configuration before creating the schedule in
Fabric. Boundaries are UTC instants, while recurring clock times use a Windows
time-zone identifier so Fabric can apply daylight-saving rules.

```{r, eval = FALSE}
daily <- fabric_job_schedule_config(
  "Daily",
  start_time = "2026-10-01T00:00:00Z",
  end_time = "2027-10-01T00:00:00Z",
  time_zone = "W. Europe Standard Time",
  times = "08:30"
)
```

Fabric supports minute-interval, daily, weekly, and monthly schedules. Weekly
schedules add `weekdays`; monthly schedules select a numbered day or an ordinal
weekday. See `?fabric_job_schedule_config` for those shapes. Use a Windows time
zone such as `"W. Europe Standard Time"`, not an IANA name such as
`"Europe/Amsterdam"`.

```{r, eval = FALSE}
weekly <- fabric_job_schedule_config(
  "Weekly",
  start_time = "2026-10-01T00:00:00Z",
  end_time = "2027-10-01T00:00:00Z",
  time_zone = "W. Europe Standard Time",
  times = "07:30",
  weekdays = c("Monday", "Thursday")
)
```

## Create, list, update, and disable

Create one schedule from a validated configuration, then list the schedules for
the item:

```{r, eval = FALSE}
schedule <- fabric_job_schedule_create(notebook, weekly, enabled = TRUE)
schedules <- fabric_job_schedules(notebook)
```

Disable or restart a schedule without rebuilding its configuration:

```{r, eval = FALSE}
disabled <- fabric_job_schedule_update(
  notebook,
  schedule,
  enabled = FALSE
)

restarted <- fabric_job_schedule_update(
  notebook,
  schedule,
  enabled = TRUE
)
```

Fabric may disable a schedule after repeated failures. Re-enable it only after
inspecting recent job history and correcting the cause.

## Delete a schedule

Deletion is permanent and never inferred from a create or update conflict. It
requires explicit confirmation:

```{r, eval = FALSE}
fabric_job_schedule_delete(notebook, schedule, confirm = TRUE)
```

Microsoft documents scheduler throttling, maximum job duration and concurrency,
schedule expiry after prolonged user inactivity, and auto-disable behavior in
the [Fabric job scheduler
guide](https://learn.microsoft.com/en-us/fabric/fundamentals/job-scheduler).
