---
title: "Working with Fabric Lakehouses and OneLake"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with Fabric Lakehouses and OneLake}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

A Fabric Lakehouse keeps ordinary files and managed tables together in
OneLake:

- `Files/` contains ordinary files such as CSV, Parquet, JSON, images, and
  scripts.
- `Tables/` contains managed Delta tables that Fabric engines can query.

Use the SQL analytics endpoint for familiar SQL queries against managed tables.
Use the Lakehouse and OneLake helpers when you want to work directly with a
table or file. This guide starts with a small SQL query, then moves from
ordinary files to managed tables and, finally, large or historical reads.

## Find and connect to a Lakehouse

Start by finding the workspace and Lakehouse by name:

```{r, eval = FALSE}
library(fabricQueryR)

workspaces <- fabric_workspaces()
matches <- Filter(
  \(x) identical(x$displayName, "Analytics workspace"),
  workspaces
)
stopifnot(length(matches) == 1L)
workspace <- matches[[1L]]
lakehouse <- workspace$lakehouses()[[1L]]

lakehouse$displayName
```

`$lakehouses()` is the workspace method for `fabric_lakehouses()`.

Fabric provides a SQL analytics endpoint for querying the managed tables under
`Tables/`. It does not query ordinary files under `Files/`. The endpoint
address is already included in `lakehouse`, so you do not need to copy it from
the Fabric portal.

The discovered `FabricLakehouse` is a read-only R6 object. For a single SQL
query, call `$sql_query()` (`fabric_sql_query()`):

```{r, eval = FALSE}
orders <- lakehouse$sql_query(
  "SELECT TOP 10 * FROM dbo.orders"
)
```

The method opens and closes the SQL connection for you. If you want to run
several commands with 'DBI', use `$sql_connect()` (`fabric_sql_connect()`):

```{r, eval = FALSE}
con <- lakehouse$sql_connect()
DBI::dbListTables(con)
DBI::dbGetQuery(con, "SELECT TOP 10 * FROM dbo.orders")
DBI::dbDisconnect(con)
```

For most work with managed tables, start with SQL. Use `$sql_query()`
(`fabric_sql_query()`) for filters, joins, and summaries, or
`$sql_read_table()` (`fabric_sql_read_table()`) to read one table without
writing SQL. Use `$sql_connect()` (`fabric_sql_connect()`) only when you want
to keep a connection open for several 'DBI' calls.

You can also read a Delta table directly through OneLake with `$read_table()`
(`fabric_lakehouse_read_table()`). This is useful for an Arrow stream or an
earlier table version. The `$onelake_*()` methods correspond to the
`fabric_onelake_*()` functions and work with ordinary files under `Files/`,
while `$write_table()` (`fabric_lakehouse_write_table()`) adds or replaces
managed table data. These operations do not need a SQL connection. Never change
the files underneath `Tables/` directly because they are part of a managed
Delta table.

## List and read ordinary files

Start by listing a small folder with `$onelake_list()`
(`fabric_onelake_list()`):

```{r, eval = FALSE}
files <- lakehouse$onelake_list(
  path = "Files/incoming"
)

files[c("path", "is_directory", "content_length")]
```

Read a supported tabular file directly into R with `$onelake_read_file()`
(`fabric_onelake_read_file()`):

```{r, eval = FALSE}
orders <- lakehouse$onelake_read_file(
  path = "Files/incoming/orders.csv"
)

head(orders)
```

Use `$onelake_metadata()` (`fabric_onelake_metadata()`) for size and other
properties without reading the contents. Use `$onelake_download()`
(`fabric_onelake_download()`) for a file type that should stay as bytes or be
saved to local disk.

## Write or upload a file

`$onelake_write_file()` (`fabric_onelake_write_file()`) turns an R or Arrow
object into CSV, Parquet, or Arrow IPC content based on the path or explicit
format:

```{r, eval = FALSE}
lakehouse$onelake_write_file(
  path = "Files/exports/orders.parquet",
  data = data.frame(
    order_id = 1:3,
    amount = c(10.5, 20, 30.25)
  )
)
```

If the file already exists on your computer, upload it with `$onelake_upload()`
(`fabric_onelake_upload()`) without first parsing it in R:

```{r, eval = FALSE}
lakehouse$onelake_upload(
  path = "Files/incoming/logo.png",
  source = "logo.png"
)
```

Both operations can replace a destination. Use a new path while learning and
check `?fabric_onelake_write_file` or `?fabric_onelake_upload` before replacing
shared data.

## Discover and read managed tables

List tables with `$tables()` (`fabric_lakehouse_tables()`):

```{r, eval = FALSE}
tables <- lakehouse$tables()
tables[c("schema", "name", "type", "format")]
```

Select a discovered row or supply the table name to `$read_table()`
(`fabric_lakehouse_read_table()`):

```{r, eval = FALSE}
table <- tables[1L, ]

rows <- lakehouse$read_table(
  table,
  columns = c("order_id", "amount"),
  limit = 100L
)
```

Column selection and row limits happen before the result is fully collected in
R. For SQL joins, grouping, or complex filters, use `$sql_query()`
(`fabric_sql_query()`) against the Lakehouse SQL analytics endpoint instead.

## Write an R object as a managed table

Use `$write_table()` (`fabric_lakehouse_write_table()`):

```{r, eval = FALSE}
result <- lakehouse$write_table(
  table = "orders_from_r",
  data = data.frame(
    order_id = 1:3,
    amount = c(10.5, 20, 30.25)
  ),
  mode = "Overwrite"
)

result$rows
result$staging_retained
```

The function stages Parquet files under a unique `Files/` path, asks Fabric to
perform a managed Delta load, waits for completion, and cleans up after
confirmed success. Fabric can infer a missing table's schema. Use Spark or a
separately schema-controlled process when inference is not suitable.

## Turn existing files into a table

When CSV or Parquet files already exist in the Lakehouse, ask Fabric to load
them directly with `$load_table()` (`fabric_lakehouse_load_table()`):

```{r, eval = FALSE}
operation <- lakehouse$load_table(
  table = "orders_from_csv",
  path = "Files/incoming/orders.csv",
  format = "Csv",
  header = TRUE,
  delimiter = ",",
  mode = "Overwrite"
)

operation <- fabric_operation_wait(operation, timeout = 900)
```

The source can be one file or a folder. Folder loads can filter by extension
and include nested folders. The source must already be inside the selected
Lakehouse's `Files/` area.

## Read a large or historical Delta table

Use `$read_table()` (`fabric_lakehouse_read_table()`) with an Arrow stream when
the selected data may not fit comfortably in memory:

```{r, eval = FALSE}
stream <- lakehouse$read_table(
  table = "large_orders",
  result = "arrow_stream"
)
reader <- arrow::as_record_batch_reader(stream)
orders <- reader$read_table()
reader$Close()
```

The stream is disk-backed and single-use. Close the Arrow reader when finished
so its staged temporary file is deleted; if you consume the 'nanoarrow' stream
directly, call `stream[["release"]]()`. Do not rely on garbage collection for
this cleanup. `version` can select an earlier Delta table version when that
history is available:

```{r, eval = FALSE}
older_rows <- lakehouse$read_table(
  table = "orders",
  version = 42L,
  limit = 100L
)
```

`fabric_onelake_read_delta_table()` is the lower-level reader for compatible
Delta paths in Lakehouses and Warehouses. Prefer the Lakehouse wrapper for a
normal discovered-table workflow.
