---
title: "Working with Fabric LakeHouses and OneLake"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with Fabric LakeHouses and OneLake}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

A Fabric Lakehouse gives you two related ways to organize data:

- `Files/` contains ordinary files such as CSV, Parquet, JSON, images, and
  scripts.
- `Tables/` contains managed Delta tables that Fabric engines can query.

An ordinary file can be copied, downloaded, or replaced. A Delta table also
has a transaction log that records valid table versions. Use table functions
for managed tables and file functions for ordinary files; never change the
files underneath `Tables/` directly.

## Find a Lakehouse

```{r, eval = FALSE}
library(fabricQueryR)

workspace <- fabric_workspaces()[["Analytics workspace"]]
lakehouse <- fabric_lakehouses(workspace)[[1L]]

lakehouse$displayName
```

Passing this discovered object to later calls avoids copying workspace IDs,
item IDs, and service addresses.

## List and read ordinary files

Start by listing a small folder:

```{r, eval = FALSE}
files <- fabric_onelake_list(
  workspace,
  lakehouse,
  path = "Files/incoming"
)

files[c("path", "is_directory", "content_length")]
```

Read a supported tabular file directly into R:

```{r, eval = FALSE}
orders <- fabric_onelake_read_file(
  workspace,
  lakehouse,
  path = "Files/incoming/orders.csv"
)

head(orders)
```

Use `fabric_onelake_metadata()` for size and other properties without reading
the contents. Use `fabric_onelake_download()` for a file type that should stay
as bytes or be saved to local disk.

## Write or upload a file

`fabric_onelake_write_file()` turns an R or Arrow object into CSV, Parquet, or
Arrow IPC content based on the path or explicit format:

```{r, eval = FALSE}
fabric_onelake_write_file(
  workspace,
  lakehouse,
  path = "Files/exports/orders.parquet",
  data = data.frame(
    order_id = 1:3,
    amount = c(10.5, 20, 30.25)
  )
)
```

If the file already exists on your computer, upload it without first parsing
it in R:

```{r, eval = FALSE}
fabric_onelake_upload(
  workspace,
  lakehouse,
  path = "Files/incoming/logo.png",
  source = "logo.png"
)
```

Both operations can replace a destination. Use a new path while learning and
check `?fabric_onelake_write_file` or `?fabric_onelake_upload` before replacing
shared data.

## Discover and read managed tables

```{r, eval = FALSE}
tables <- fabric_lakehouse_tables(lakehouse)
tables[c("schema", "name", "type", "format")]
```

Select a discovered row or supply the table name:

```{r, eval = FALSE}
table <- tables[1L, ]

rows <- fabric_lakehouse_read_table(
  lakehouse,
  table,
  columns = c("order_id", "amount"),
  limit = 100L
)
```

Column selection and row limits happen before the result is fully collected in
R. For SQL joins, grouping, or complex filters, use `fabric_sql_query()`
against the Lakehouse SQL analytics endpoint instead.

## Write an R object as a managed table

```{r, eval = FALSE}
result <- fabric_lakehouse_write_table(
  lakehouse,
  table = "orders_from_r",
  data = data.frame(
    order_id = 1:3,
    amount = c(10.5, 20, 30.25)
  ),
  mode = "Overwrite"
)

result$rows
result$staging_retained
```

The function stages Parquet files under a unique `Files/` path, asks Fabric to
perform a managed Delta load, waits for completion, and cleans up after
confirmed success. Fabric can infer a missing table's schema. Use Spark or a
separately schema-controlled process when inference is not suitable.

## Turn existing files into a table

When CSV or Parquet files already exist in the Lakehouse, ask Fabric to load
them directly:

```{r, eval = FALSE}
operation <- fabric_lakehouse_load_table(
  lakehouse,
  table = "orders_from_csv",
  path = "Files/incoming/orders.csv",
  format = "Csv",
  header = TRUE,
  delimiter = ",",
  mode = "Overwrite"
)

operation <- fabric_operation_wait(operation, timeout = 900)
```

The source can be one file or a folder. Folder loads can filter by extension
and include nested folders. The source must already be inside the selected
Lakehouse's `Files/` area.

## Read a large or historical Delta table

Use an Arrow stream when the selected data may not fit comfortably in memory:

```{r, eval = FALSE}
stream <- fabric_lakehouse_read_table(
  lakehouse,
  table = "large_orders",
  result = "arrow_stream"
)
reader <- arrow::as_record_batch_reader(stream)
```

The stream is disk-backed and single-use. `version` can select an earlier Delta
table version when that history is still available:

```{r, eval = FALSE}
older_rows <- fabric_lakehouse_read_table(
  lakehouse,
  table = "orders",
  version = 42L,
  limit = 100L
)
```

`fabric_onelake_read_delta_table()` is the lower-level reader for compatible
Delta paths in Lakehouses and Warehouses. Prefer the Lakehouse wrapper for a
normal discovered-table workflow.

## Use shortcuts instead of copying data

A OneLake shortcut makes data stored elsewhere appear at a path in the current
item. It is a link, not a second copy:

```{r, eval = FALSE}
lakehouses <- fabric_lakehouses(workspace)
destination <- lakehouses[[1L]]
source <- lakehouses[[2L]]

shortcut <- fabric_onelake_shortcut_create(
  destination,
  path = "Files",
  name = "shared-orders",
  target = source,
  target_path = "Tables/orders"
)
```

List shortcuts with `fabric_onelake_shortcuts()`. Deleting a shortcut requires
`confirm = TRUE` and removes only the link; it does not delete the target data.

## Treat deletion as a separate step

`fabric_onelake_delete()` can permanently remove a file or directory. Inspect
the exact path with `fabric_onelake_list()` or `fabric_onelake_metadata()`
first. Keep destructive cleanup separate from read, write, and verification
code so that a failed earlier step cannot accidentally broaden what is removed.
