---
title: "Bring Fabric data into R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bring Fabric data into R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Reading data means selecting data stored or served by Fabric and returning it
to your local R session. For ordinary analysis, the result is usually a
*tibble* (a modern R data frame). It's also possible to get an Arrow stream
for larger-than-memory data.

The best method depends mainly on where the data lives and whether you want
whole tables or a filtered result. Start with a small, filtered read. Move to
streaming or Spark only when the data is too large or the transformation needs
distributed compute.

## Choose a reading method

| Your source or goal | Start with | Why |
|---|---|---|
| Warehouse, SQL Database, or a SQL-shaped Lakehouse query | `fabric_sql_query()` | Filter and summarize on the server with familiar SQL |
| One Lakehouse Delta table | `fabric_lakehouse_read_table()` | Read a table without writing SQL |
| One Warehouse table | `fabric_warehouse_read_table()` | Read a table by name without writing SQL |
| Eventhouse table or KQL result | `fabric_kql_read_table()` or `fabric_kql_query()` | Use the Eventhouse query engine |
| Power BI/Fabric semantic model | `fabric_pbi_dax_query()` | Use model relationships and measures |
| CSV, Parquet, or Arrow file in OneLake | `fabric_onelake_read_file()` | Read the file itself |
| API for GraphQL | `fabric_graphql_query()` | Request the fields exposed by the API |
| A transformation that genuinely needs Spark | `fabric_livy_query()` | Run distributed code in Fabric |

SQL is a good default for relational data because Fabric does the filtering and
R receives only the rows you need. A direct table reader is simpler when you
want one table and do not need joins or grouping.

## Discover the data source once

The examples use discovered objects. They carry the IDs and connection details
needed by the read functions:

```{r, eval = FALSE}
library(fabricQueryR)

workspaces <- fabric_workspaces()
matches <- Filter(
  \(x) identical(x$displayName, "Analytics workspace"),
  workspaces
)
stopifnot(length(matches) == 1L)
workspace <- matches[[1L]]
lakehouse <- fabric_lakehouses(workspace)[[1L]]
warehouse <- fabric_warehouses(workspace)[[1L]]
kql_database <- fabric_kql_databases(workspace)[[1L]]
model <- fabric_semantic_models(workspace)[[1L]]
```

Your workspace does not need to contain every item type. Discover only the
source used by your workflow.

## Read with SQL

Use `fabric_sql_query()` for one read-only `SELECT` query. It opens and closes
the connection for you and returns a tibble:

```{r, eval = FALSE}
recent_orders <- fabric_sql_query(
  warehouse,
  sql = paste(
    "SELECT TOP 100 order_id, order_date, amount",
    "FROM dbo.orders",
    "WHERE order_date >= ?",
    "ORDER BY order_date DESC"
  ),
  params = list(as.Date("2026-01-01"))
)

head(recent_orders)
```

Put changing values in `params` rather than pasting them into the SQL text.
This handles quoting safely. Use `fabric_sql_connect()` and normal `DBI`
functions when several queries should share one connection:

```{r, eval = FALSE}
con <- fabric_sql_connect(warehouse)
on.exit(DBI::dbDisconnect(con), add = TRUE)

DBI::dbListTables(con)
orders <- DBI::dbGetQuery(con, "SELECT TOP 100 * FROM dbo.orders")
```

The default SQL backend uses Microsoft ODBC Driver 18 for SQL Server. See
`?fabric_sql_connect` for the ODBC and ADBC setup choices.

## Read one table without writing a query

Lakehouse and Warehouse table readers accept a table name and can limit the
data before it enters R:

```{r, eval = FALSE}
lakehouse_rows <- fabric_lakehouse_read_table(
  lakehouse,
  table = "orders",
  columns = c("order_id", "order_date", "amount"),
  limit = 100L
)

warehouse_rows <- fabric_warehouse_read_table(
  warehouse,
  table = "orders",
  schema = "dbo",
  limit = 100L
)
```

Use `fabric_lakehouse_tables(lakehouse)` when you are unsure which Lakehouse
tables or schemas are available.

## Read Eventhouse data

An Eventhouse is optimized for event, log, and time-series data. Its query
language is KQL. Read a whole table by name for a simple start:

```{r, eval = FALSE}
events <- fabric_kql_read_table(
  kql_database,
  table = "Events",
  limit = 100L
)
```

Use a KQL query when Fabric should filter or summarize the events first:

```{r, eval = FALSE}
daily_events <- fabric_kql_query(
  kql_database,
  query = paste(
    "Events",
    "| where observed_at >= ago(7d)",
    "| summarize event_count = count() by bin(observed_at, 1d)",
    "| order by observed_at asc"
  )
)
```

## Query a semantic model with DAX

A semantic model is a dataset ready for reporting, commonly used in PowerBi.
Semantic models can be queried with DAX (Data Analysis Expressions):

```{r, eval = FALSE}
sales_by_region <- fabric_pbi_dax_query(
  model,
  dax = paste(
    "EVALUATE",
    "SUMMARIZECOLUMNS(",
    "  'Region'[Region],",
    "  \"Total Sales\", [Total Sales]",
    ")"
  )
)
```

The [Semantic Models vignette](semantic-model-refresh.html) shows more
things you can do with semantic models, like refreshing their data.

## Read a OneLake file

Use the file reader when the file itself is the data contract. Paths in a
Lakehouse usually begin with `Files/`:

```{r, eval = FALSE}
orders_file <- fabric_onelake_read_file(
  workspace,
  lakehouse,
  path = "Files/exports/orders.parquet"
)
```

CSV, Parquet, and Arrow IPC files can become R or Arrow objects. Other file
types can be downloaded as raw bytes or to disk. Use a Lakehouse table reader,
not a file reader, for managed data below `Tables/`.

## Read through GraphQL or Spark

GraphQL is useful when a Fabric API item exposes a purpose-built selection of
fields:

```{r, eval = FALSE}
api <- fabric_graphql_apis(workspace)[[1L]]

response <- fabric_graphql_query(
  api,
  query = "{ products { items { id name category } } }"
)
products <- response$data$products$items
```

Use [Working with GraphQL](graphql-schema-and-rows.html) for schema inspection
and pagination.

Spark is the later choice for distributed transformations, Spark-only data
formats, or logic already written for Spark:

```{r, eval = FALSE}
result <- fabric_livy_query(
  lakehouse,
  kind = "sql",
  code = "SELECT category, count(*) AS n FROM orders GROUP BY category"
)
counts <- result$output$parsed
```

Spark has startup cost. Prefer SQL or a direct reader for a small, ordinary
table read.

## Working with large data (Arrow streams)

If you work with large volumes of data (larger than your device's working memory),
you may need to stream the data into R rather than reading it all at once.

Several readers accept `result = "arrow_stream"`. This can be used with the
'arrow' R package to read the data in batches:

```{r, eval = FALSE}
stream <- fabric_lakehouse_read_table(
  lakehouse,
  table = "large_orders",
  result = "arrow_stream"
)

reader <- arrow::as_record_batch_reader(stream)
```

See the [Arrow R package](https://arrow.apache.org/docs/r/) for more information 
on working with Arrow streams and record batches.
