---
title: "Query and refresh a semantic model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Query and refresh a semantic model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

A semantic model is the report-ready data behind Power BI and Fabric reports.
It combines tables with relationships, measures, and business calculations.
From R, you can query that model with DAX and refresh it after its source data
changes.

Start by querying a small result. Refresh and monitoring come next, followed by
the more specialized controls used for production models.

This guide uses a discovered model so its workspace and dataset IDs travel
together:

```{r}
library(fabricQueryR)

model <- fabric_semantic_models("Analytics workspace")[[1]]
```

## Query the model with DAX

DAX is the query and calculation language used by semantic models. A query
normally begins with `EVALUATE` and returns a table. If you are new to DAX,
create and test a query in the model's **DAX query view** in Fabric, then use the
same text from R:

```{r}
rows <- fabric_pbi_dax_query(
  model,
  dax = "EVALUATE TOPN(100, 'Customers')"
)

head(rows)
```

Query the semantic model when you want its measures and relationships, or need
the same business definition used by a report. Query the Warehouse or
Lakehouse instead when you need the underlying source rows and SQL is a better
fit.

For example, a DAX measure can be grouped by a model column:

```{r}
sales <- fabric_pbi_dax_query(
  model,
  dax = paste(
    "EVALUATE",
    "SUMMARIZECOLUMNS(",
    "  'Region'[Region],",
    "  \"Total Sales\", [Total Sales]",
    ")"
  )
)
```

The normal JSON mode returns one tibble and is the simplest option. The later
`api = "arrow"` mode is useful for richer types, multiple result tables, or a
stream, but requires a capacity-backed model and additional tenant settings.

## Check access before refreshing

Querying requires Read and Build permission on the model and the tenant's
**Semantic Model Execute Queries REST API** setting. Starting or cancelling a
refresh requires `Dataset.ReadWrite.All` and Write permission on the model.
Status and history use `Dataset.Read.All` or `Dataset.ReadWrite.All`, although
Microsoft also requires model Write permission for refresh history.

Service principals are supported when tenant settings and model or workspace
access allow them. Do not set an email notification option for a
service-principal refresh request.

## Refresh after an upstream update

A standard refresh processes the complete model using Power BI defaults:

```{r}
refresh <- fabric_pbi_refresh(model)
completed <- fabric_pbi_refresh_wait(refresh, timeout = 1800)

completed$state
completed$start_time
completed$end_time
```

The returned handle stores its target and in-process sign-in context. It can be
passed to status, wait, or cancel without copying IDs. Credentials are not
persisted when a handle is serialized; authenticate again after restoring one
in another R process.

For example, a data-load workflow can refresh only after the source update has
committed successfully:

```{r}
load <- fabric_lakehouse_write_table(
  lakehouse,
  "dbo.Sales",
  new_sales,
  mode = "overwrite",
  wait = TRUE
)

refresh <- fabric_pbi_refresh(model)
completed <- fabric_pbi_refresh_wait(
  refresh,
  timeout = 1800,
  cancel_on_timeout = TRUE
)
```

Client-side `timeout` in `fabric_pbi_refresh_wait()` only bounds how long R
waits. Without `cancel_on_timeout = TRUE`, the Power BI refresh keeps running.

## Use enhanced refresh controls

Enhanced refresh requires Power BI Premium, Premium per user, Embedded, or
Fabric capacity. Supplying an enhanced option selects enhanced mode
automatically, or set `mode = "enhanced"` explicitly:

```{r}
refresh <- fabric_pbi_refresh(
  model,
  mode = "enhanced",
  type = "Full",
  objects = list(
    list(table = "Sales", partition = "2026"),
    list(table = "Calendar")
  ),
  commit_mode = "Transactional",
  max_parallelism = 4L,
  retry_count = 1L,
  timeout = "02:00:00"
)

completed <- fabric_pbi_refresh_wait(refresh, timeout = 5 * 60 * 60)
```

The refresh submission `timeout` is an `HH:MM:SS` limit for each attempt. Each
retry gets a new attempt limit, but Power BI limits the whole operation to 24
hours. The wait function's numeric `timeout` remains a separate client limit.

Use `Transactional` to preserve the previous model unless the entire refresh
succeeds. `PartialBatch` commits each processing command separately, so a
failure can leave only part of the new data loaded or a table empty. It also
cannot be combined with `apply_refresh_policy = TRUE`.

Power BI permits only one active refresh per semantic model. A second request
is rejected rather than queued behind the first.

## Diagnose attempts and failures

Execution details retain the service's general and extended states. The
normalized `state` distinguishes `Queued`, `InProgress`, `Completed`,
`CompletedWithWarnings`, `Failed`, `TimedOut`, `Cancelled`, and `Disabled`.

```{r}
status <- fabric_pbi_refresh_status(refresh)

status$state
status$attempts
status$messages
status$service_error
status$objects
status$details_url
```

Each attempt retains Data or Query phase information, timestamps, execution
metrics, and a parsed service exception. The complete response remains in
`raw`, so new Microsoft fields are still inspectable.

Power BI can finish successfully with non-fatal warnings. Microsoft currently
documents that the history and execution-detail REST APIs do not include those
warnings. If the service returns warning messages, fabricQueryR reports
`CompletedWithWarnings`; otherwise open `details_url` to check the Fabric
refresh-detail page.

By default, wait raises typed R conditions for service failures, cancellation,
service timeouts, and disabled refreshes. Keep the terminal detail instead when
building a monitoring table:

```{r}
result <- fabric_pbi_refresh_wait(
  refresh,
  error_on_failure = FALSE
)

if (result$state != "Completed") result$details_url
```

## Inspect history or cancel work

Power BI retains between 20 and 60 recent history entries depending on their
age. OneDrive synchronization history is separate and is not returned here.

```{r}
history <- fabric_pbi_refresh_history(model, top = 10L)

history[[1]]$refresh_type
history[[1]]$state
history[[1]]$attempts

# Refresh an old history entry from its request ID and stored model context
latest <- fabric_pbi_refresh_status(history[[1]])
```

Cancel a running Import or Composite model refresh when the model is on
Premium, PPU, Embedded, or Fabric capacity. Microsoft requires Contributor,
Member, or Admin access to the workspace:

```{r}
fabric_pbi_refresh_cancel(refresh)
```

## Capacity and Direct Lake behavior

Shared capacity supports standard refresh only and permits at most eight
scheduled and API-triggered refresh requests per day. Capacity-backed models
have no fixed API-refresh count, but work can queue or throttle as capacity
resources become busy. A refresh fails if capacity throttling continues beyond
the service limit.

Direct Lake refresh is different from Import refresh. It frames the current
Delta table versions so the model has a new point-in-time baseline; it does not
copy all OneLake data into the model. Automatic Direct Lake updates are enabled
by default, so explicit refresh is commonly unnecessary. Use it when automatic
updates are disabled, after programmatic table changes, or when a controlled
consistent frame is part of the workflow.

See Microsoft's documentation for the current contracts and limits:

- [Refresh Dataset API](https://learn.microsoft.com/en-us/rest/api/power-bi/datasets/refresh-dataset-in-group)
- [Enhanced refresh](https://learn.microsoft.com/en-us/power-bi/connect-data/asynchronous-refresh)
- [Refresh execution details](https://learn.microsoft.com/en-us/rest/api/power-bi/datasets/get-refresh-execution-details-in-group)
- [Data refresh and capacity limits](https://learn.microsoft.com/en-us/power-bi/connect-data/refresh-data)
- [How Direct Lake works](https://learn.microsoft.com/en-us/fabric/fundamentals/direct-lake-how-it-works)
