---
title: "Working with Semantic Models (DAX queries)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with Semantic Models (DAX queries)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

A semantic model is the report-ready data behind Power BI and Fabric reports.
It combines tables with relationships, measures, and business calculations.
From R, you can query that model with DAX and refresh it after its source data
changes.

Start by querying a small result. Refresh and monitoring come next, followed by
the more specialized controls used for production models.

This guide uses a discovered model so its workspace and dataset IDs travel
together:

```{r, eval = FALSE}
library(fabricQueryR)

model <- fabric_semantic_models("Analytics workspace")[[1]]
```

## Query the model with DAX

DAX is the query and calculation language used by semantic models. A query
normally begins with `EVALUATE` and returns a table. If you are new to DAX,
create and test a query in the model's *DAX query view* in Fabric, then use the
same text from R:

```{r, eval = FALSE}
rows <- fabric_pbi_dax_query(
  model,
  dax = "EVALUATE TOPN(100, 'Customers')"
)

head(rows)
```

Query the semantic model when you want its measures and relationships, or need
the same business definition used by a report. Query the Warehouse or
Lakehouse instead when you need the underlying source rows and SQL is a better
fit.

For example, a DAX measure can be grouped by a model column:

```{r, eval = FALSE}
sales <- fabric_pbi_dax_query(
  model,
  dax = paste(
    "EVALUATE",
    "SUMMARIZECOLUMNS(",
    "  'Region'[Region],",
    "  \"Total Sales\", [Total Sales]",
    ")"
  )
)
```

The normal JSON mode returns one tibble and is the simplest option. The later
`api = "arrow"` mode is useful for richer types, multiple result tables, or a
stream, but requires a capacity-backed model and additional tenant settings.

## Check access before refreshing

Querying normally requires Read and Build permission on the model and the
tenant setting that enables Execute Queries. Refreshing requires Write access.
If a query succeeds but a refresh does not, ask the model owner or Fabric
administrator to check those permissions and tenant settings.

## Refresh after an upstream update

A standard refresh processes the complete model using Power BI defaults:

```{r, eval = FALSE}
refresh <- fabric_pbi_refresh(model)
completed <- fabric_pbi_refresh_wait(refresh, timeout = 1800)

completed$state
completed$start_time
completed$end_time
```

Pass the returned handle to status, wait, or cancel functions without copying
IDs. If you save it and restore it in another R process, authenticate again.

For example, a data-load workflow can refresh only after the source update has
committed successfully:

```{r, eval = FALSE}
load <- fabric_lakehouse_write_table(
  lakehouse,
  table = "Sales",
  data = new_sales,
  schema = "dbo",
  mode = "overwrite"
)

refresh <- fabric_pbi_refresh(model)
completed <- fabric_pbi_refresh_wait(
  refresh,
  timeout = 1800,
  cancel_on_timeout = TRUE
)
```

Client-side `timeout` in `fabric_pbi_refresh_wait()` only bounds how long R
waits. Without `cancel_on_timeout = TRUE`, the Power BI refresh keeps running.

## Use enhanced refresh controls

Enhanced refresh requires Power BI Premium, Premium per user, Embedded, or
Fabric capacity. Supplying an enhanced option selects enhanced mode
automatically, or set `mode = "enhanced"` explicitly:

```{r, eval = FALSE}
refresh <- fabric_pbi_refresh(
  model,
  mode = "enhanced",
  type = "Full",
  objects = list(
    list(table = "Sales", partition = "2026"),
    list(table = "Calendar")
  ),
  commit_mode = "Transactional",
  max_parallelism = 4L,
  retry_count = 1L,
  timeout = "02:00:00"
)

completed <- fabric_pbi_refresh_wait(refresh, timeout = 5 * 60 * 60)
```

The submission `timeout` controls an individual Power BI attempt; the numeric
timeout in `fabric_pbi_refresh_wait()` controls how long R waits. Use
`"Transactional"` when the previous model should remain available unless the
complete refresh succeeds. See `?fabric_pbi_refresh` before using partial-batch
commits or refresh-policy options.

## Diagnose attempts and failures

Use the status helper when a refresh takes longer than expected or fails:

```{r, eval = FALSE}
status <- fabric_pbi_refresh_status(refresh)

status$state
status$attempts
status$messages
status$service_error
status$objects
status$details_url
```

The normalized state distinguishes work that is queued, running, completed,
completed with warnings, failed, timed out, cancelled, or disabled. Attempt
details and `details_url` help diagnose the service response; the complete
response remains available in `raw`.

By default, wait raises typed R conditions for service failures, cancellation,
service timeouts, and disabled refreshes. Keep the terminal detail instead when
building a monitoring table:

```{r, eval = FALSE}
result <- fabric_pbi_refresh_wait(
  refresh,
  error_on_failure = FALSE
)

if (result$state != "Completed") result$details_url
```

## Inspect history

Inspect recent refreshes when you need to compare duration or failure patterns:

```{r, eval = FALSE}
history <- fabric_pbi_refresh_history(model, top = 10L)

history[[1]]$refresh_type
history[[1]]$state
history[[1]]$attempts

# Refresh an old history entry from its request ID and stored model context
latest <- fabric_pbi_refresh_status(history[[1]])
```

## More information

See Microsoft's documentation for more information:
- [Refresh Dataset API](https://learn.microsoft.com/en-us/rest/api/power-bi/datasets/refresh-dataset-in-group)
- [Enhanced refresh](https://learn.microsoft.com/en-us/power-bi/connect-data/asynchronous-refresh)
- [Refresh execution details](https://learn.microsoft.com/en-us/rest/api/power-bi/datasets/get-refresh-execution-details-in-group)
- [Data refresh and capacity limits](https://learn.microsoft.com/en-us/power-bi/connect-data/refresh-data)
- [How Direct Lake works](https://learn.microsoft.com/en-us/fabric/fundamentals/direct-lake-how-it-works)
