---
title: "Shiny apps using each visitor's Fabric access"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Shiny apps using each visitor's Fabric access}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

fabricQueryR lets Shiny apps query Microsoft Fabric using each signed-in user's
permissions. Visitors sign in with a Microsoft account, and requests to Fabric
use that account's access to the data.

You can use Shiny functions such as `reactive()` and `renderTable()`,
together with fabricQueryR's SQL and GraphQL functions. The shinyOAuth package
handles the Microsoft sign-in process. You add a small amount of configuration
and connect it to your app's UI and server.

This vignette starts with an app that lists the tables a user can see in a
Fabric SQL database. It then shows how to work with a Warehouse, load data on
demand, and adapt an app that already uses shinyOAuth.

## Setup

### Install the packages

This integration currently requires the development versions of fabricQueryR
and shinyOAuth. If you have both source checkouts, install shinyOAuth first:

```{r, eval=FALSE}
devtools::install("../shinyoauth")
devtools::install(".")
```

For the SQL example, you also need shiny, DBI, odbc, and Microsoft
ODBC Driver 18 for SQL Server. The driver is installed separately from the R
packages, on the computer that runs the app.

### Arrange Microsoft sign-in

Your app needs an application registration in Microsoft Entra, the service that
manages your organization's Microsoft sign-in. Your Fabric or Microsoft 365
administrator can help set this up.

For the SQL example, ask for a Web application registration with the Azure
SQL Database delegated permission `user_impersonation`. A delegated permission
allows the app to make requests using the signed-in person's access. Users still
need their own permission to connect to the database and read its data.

The registration needs a redirect URI: the address where Microsoft sends
the visitor after sign-in. For local development, use `http://localhost:8100/`.
The address in the registration and the app must match. A deployed app uses its
public HTTPS address instead.

You will need three values from the registration: the tenant ID, which
identifies your organization's directory; the client ID, which identifies
your app; and a client secret, which the app uses during sign-in. Complete
any administrator consent your organization requires before testing the app.

### Save the connection settings

The examples read settings with `Sys.getenv()`. For local development, you can
put them in your user `.Renviron` file and restart R:

```text
ENTRA_TENANT_ID=your-directory-id
ENTRA_CLIENT_ID=your-application-id
ENTRA_CLIENT_SECRET=your-client-secret
FABRIC_SQL_SERVER=your-sql-endpoint-hostname
FABRIC_SQL_DATABASE=your-database-name
```

Copy the SQL server and database name from the connection details for your
Fabric Warehouse, SQL Database or Lakehouse SQL analytics endpoint. Keep the
client secret out of your app code and version control. For a deployed app,
use the hosting service's environment-variable or secret settings.

## Try the example app

The package includes a complete app with sign-in, sign-out and a table-name
filter. After setting the values above, run:

```{r, eval=FALSE}
example_dir <- system.file("examples", "shiny-fabric", package = "fabricQueryR")
shiny::runApp(example_dir, host = "127.0.0.1", port = 8100)
```

Open `http://localhost:8100/`, sign in, and click Show tables. The app lists
tables visible to your account; an empty result can be valid if the account
cannot see any tables. It does not create or change data.

## Build your own app

Here is a smaller app showing the three pieces you add to a Shiny application:

- `fabric_shiny_config()` holds the sign-in settings and selects the services
  the app will use. Create it once, outside `server()`.
- `fabric_shiny_ui()` wraps your usual UI so the app can complete sign-in.
- `fabric_shiny_server()` creates helpers for the current visitor. Call it
  inside `server()` so each visitor has their own connection.

```{r, eval=FALSE}
library(shiny)
library(fabricQueryR)

config <- fabric_shiny_config(
  tenant_id = Sys.getenv("ENTRA_TENANT_ID"),
  client_id = Sys.getenv("ENTRA_CLIENT_ID"),
  client_secret = Sys.getenv("ENTRA_CLIENT_SECRET"),
  redirect_uri = "http://localhost:8100/",
  services = "sql"
)
ui <- fabric_shiny_ui(
  fluidPage(actionButton("login", "Sign in"), tableOutput("tables")),
  "fabric", config
)
server <- function(input, output, session) {
  fabric <- fabric_shiny_server("fabric", config)
  observeEvent(input$login, fabric$login(), ignoreInit = TRUE)
  output$tables <- renderTable({
    req(fabric$ready("sql"))
    fabric_sql_query(
      Sys.getenv("FABRIC_SQL_SERVER"),
      "SELECT TABLE_SCHEMA, TABLE_NAME FROM INFORMATION_SCHEMA.TABLES",
      database = Sys.getenv("FABRIC_SQL_DATABASE"),
      token = fabric$token_provider(), numeric_policy = "driver"
    )
  })
}
shinyApp(ui, server, uiPattern = ".*")
```

The Sign in button calls `fabric$login()`. Inside `renderTable()`,
`req(fabric$ready("sql"))` waits until SQL access is ready to try. Passing
`token = fabric$token_provider()` tells the query to use the current visitor's
sign-in. Fabric still checks whether that person can access the requested data.

To adapt this example, replace the SQL with your own read query and add the
inputs and outputs your app needs. Use `params` for values supplied by
inputs, as in the Warehouse example below.

## Discover a Warehouse and reuse its methods

If you know a Warehouse's workspace ID and item ID, you can look it up and use
its usual fabricQueryR methods. Change the configuration to
`services = c("fabric", "sql")`: `"fabric"` lets the app look up the item, and
`"sql"` lets it query the data.

The example below assumes an existing `dbo.orders` table and a Shiny input named
`region`. It uses the ADBC backend, which needs its corresponding driver
installed. Add this code inside your server function:

```{r, eval=FALSE}
warehouse <- reactive({
  req(fabric$ready("fabric"))
  fabric$item(
    workspace = Sys.getenv("FABRIC_WORKSPACE_ID"),
    item = Sys.getenv("FABRIC_WAREHOUSE_ID"),
    type = "Warehouse"
  )
})
output$orders <- renderTable({
  req(fabric$ready("sql"))
  warehouse()$sql_query(
    "SELECT TOP (100) order_id, region FROM dbo.orders WHERE region = ?",
    params = list(input$region), backend = "adbc"
  )
})
```

The returned Warehouse object remembers the visitor's connection, so its
`$sql_query()` method does not need a separate `token` argument. You can also use
`fabric$workspaces()` to list available workspaces. Looking up an item by its
IDs avoids needing that list first.

When calling standalone functions such as `fabric_sql_query()`, always include
`token = fabric$token_provider()`, even if you pass a Warehouse object as the
target. Without it, the function uses the package's usual AzureAuth sign-in.

## Use GraphQL

An existing Fabric API for GraphQL is another way to query data, without a SQL
driver. Configure `services = "graphql"` and pass the visitor's connection to
`fabric_graphql_query()` inside your server code:

```{r, eval=FALSE}
result <- fabric_graphql_query(
  api = Sys.getenv("FABRIC_GRAPHQL_ENDPOINT"),
  query = configured_query_document,
  variables = list(region = input$region),
  token = fabric$token_provider()
)
```

Replace `configured_query_document` with a query for your API's schema. If the
underlying data source should use each visitor's permissions, check that the
API's source connection uses the appropriate single sign-on settings. A source
connection using saved credentials can give different access. See Microsoft's
[GraphQL connection guidance](https://learn.microsoft.com/en-us/fabric/data-engineering/connect-apps-api-graphql).

## Sign out and load data on demand

The full example app includes buttons for `fabric$logout()` and
`fabric$reauthorize()`, which starts sign-in again. A sign-out should also clear
any data you have stored in the app for that visitor.

This matters when a Load button runs a query: signing out does not click that
button again, so its previous result can remain in a reactive value.
`fabric$generation()` identifies the current sign-in. Store it with the result
and check it before displaying data. It changes when the visitor signs out or
signs in again, while routine credential renewal leaves it unchanged.

For example, with a Load button and the `warehouse()` reactive defined above:

```{r, eval=FALSE}
rows <- reactiveVal(NULL)
observeEvent(fabric$generation(), rows(NULL), ignoreNULL = FALSE, priority = 50)
observeEvent(input$load, {
  req(fabric$ready("sql"))
  generation <- fabric$generation()
  data <- warehouse()$sql_query(configured_read_query)
  if (identical(generation, fabric$generation())) {
    rows(list(generation = generation, data = data))
  }
}, ignoreInit = TRUE)
output$rows <- renderTable({
  req(fabric$ready("sql"))
  value <- req(rows())
  req(identical(value$generation, fabric$generation()))
  value$data
})
```

Here, `configured_read_query` is your SQL read query. Keep `rows` and other
user-specific data inside `server()`. Avoid `req(..., cancelOutput = TRUE)` for
these checks, because it keeps the previous output visible.

## Advanced configuration

The following details are useful when adapting an existing authentication setup,
troubleshooting access, or preparing an app for deployment.

### Choosing permissions

`services` selects the kinds of Fabric connection your app needs. The SQL
example uses only `"sql"`; adding `"fabric"` requests permissions to look up
workspaces and items. Its defaults are `Workspace.Read.All` and `Item.Read.All`.
You can request a narrower set with, for example,
`scopes = list(fabric = "Item.Read.All")` when using known item IDs. GraphQL
uses the Power BI service permission `GraphQLApi.Execute.All`.

The package fills in the full scope URLs, including SQL's
`https://database.windows.net//user_impersonation`. You normally do not need to
construct these yourself. See [Fabric data-plane authentication](https://learn.microsoft.com/en-us/fabric/workload-development-kit/fabric-data-plane)
and [Fabric REST scopes](https://learn.microsoft.com/en-us/rest/api/fabric/articles/scopes)
when choosing permissions for a registration.

Sign-in also requests `openid`, `profile` and `offline_access` to identify the
visitor and renew credentials. Identity comes from a validated ID token; Graph
UserInfo and `User.Read` are not needed just to display the visitor's name.
The client secret authenticates the app during sign-in, while data requests use
the visitor's delegated access. This follows Microsoft's
[authorization code flow](https://learn.microsoft.com/en-us/entra/identity-platform/v2-oauth2-auth-code-flow).

### Using an existing shinyOAuth app

If your app already handles sign-in with shinyOAuth, use
`fabric_shiny_token_provider()` with its existing connection. You do not need a
second login UI. The shinyOAuth client must already declare a token target for
the service you want to use. In this example, its SQL target is named `db`:

```{r, eval=FALSE}
provider <- reactive({
  connection <- req(auth$connection())
  fabric_shiny_token_provider(connection, services = "sql", targets = c(sql = "db"))
})
output$tables <- renderTable({
  fabric_sql_query(
    configured_server, "SELECT TABLE_NAME FROM INFORMATION_SCHEMA.TABLES",
    database = configured_database, token = provider(), numeric_policy = "driver"
  )
})
```

For several managed authorizations, select one with
`auth$connection(selected_id)` and follow shinyOAuth's ownership and activity
requirements. The provider stays tied to that connection. Configure the needed
targets before sign-in; wrapping a Graph-only connection does not add SQL access.

### Diagnosing connection problems

`fabric$ready("sql")` checks the permissions recorded for the connection. The
query itself can still fail because of network problems or database permissions.
Use `fabric$status()` to inspect connection state and
`fabric$prepare("sql")` to retry preparing SQL access. A problem with one
service can leave the others available.

For application error handling, `shinyOAuth_access_error` provides a
`context$reason` field. `interaction_required` calls for an explicit sign-in
again; `insufficient_scope` calls for checking permissions and consent.
`refresh_pending` means a synchronous call cannot yet get the renewed
credential. SQL may wrap these errors in the parent of a
`fabric_sql_authentication_error`. Show users a useful message without exposing
raw credentials or provider diagnostics.

Routine token refresh does not change `generation()` or rerun queries that
depend on it. HTTP calls retain fabricQueryR's retry behavior: one 401 response
can trigger token renewal and a request retry even with `idempotent = FALSE`.
This is separate from transient network/service retries. `prepare()` and
`reauthorize()` themselves do not replay data operations.

### Connections and deployment

`fabric_sql_query()` opens and closes its own connection. If you use
`fabric_sql_connect()` or `$sql_connect()` directly, close it with
`DBI::dbDisconnect()`, usually through `on.exit()` in a helper function. Signing
out does not close an already open DBI connection. Use `backend = "adbc"` with
its driver when exact numeric or Arrow results are needed.

For deployment, use one R process and an exact registered HTTPS callback URL.
The `uiPattern = ".*"` setting lets callback subpaths reach the app. Check your
hosting service's routing, reverse-proxy and cookie settings using shinyOAuth's
deployment guidance. The default local sign-in lifetime is eight hours; adjust
`reauth_after_seconds` for your app. App logout does not sign the visitor out of
Microsoft everywhere.

Standard Microsoft service hosts are allowed. A trusted gateway needs its exact
hostname in `endpoint_hosts`, configured by the app author. HTTP redirects are
rejected, and later request destinations are checked too. These checks protect
normal package use; they do not isolate untrusted R code.
