---
title: "Shiny apps using each visitor's Fabric access"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Shiny apps using each visitor's Fabric access}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`fabric_shiny_config()`, `fabric_shiny_ui()` and `fabric_shiny_server()` connect
an ordinary Shiny app to the visitor's delegated Microsoft Fabric access.
shinyOAuth handles sign-in, validated identity, token targets and refresh.
fabricQueryR supplies resource scopes, endpoint checks and its existing query
and discovery interfaces.

The initial integration supports Fabric discovery, SQL and GraphQL in one fixed
tenant, with one authorization per Shiny session and synchronous calls. Shiny and
shinyOAuth are optional dependencies. Install a shinyOAuth build with
`oauth_client(token_targets = ...)`, module `auth$connection()` and
`connection$access_token(target = ...)`. For the development checkouts:

```{r, eval=FALSE}
devtools::install("../shinyoauth")
devtools::install(".")
```

Local tests exercise the real module's ownership and target acquisition code
with synthetic token responses. Browser sign-in and restricted-user Fabric
acceptance remain pending. In particular, these tests do not establish your
registration's consent, tenant policies, SQL grants, GraphQL SSO or row filtering.

## Register the application

1. Create an Entra application for the fixed directory containing the Fabric
   data. Register an exact **Web** redirect URI, for example
   `http://localhost:8100/` during development and an HTTPS URL in deployment.
2. Configure the delegated permissions for the services you use. Obtain tenant
   administrator consent where required. The app requests OIDC `openid` and
   `profile`, plus `offline_access` for refresh; it does not require Graph
   UserInfo or `User.Read` just to display identity.
3. Store the application ID, tenant ID and client secret in the server's secret
   configuration. Never put a secret in app UI, a URL, a committed file or logs.
4. Grant the users access to the intended Fabric items and underlying data.
   App consent does not grant workspace, item or database access.

| Service | Delegated permission defaults | Additional authorization |
| --- | --- | --- |
| `fabric` | `Workspace.Read.All`, `Item.Read.All` | User access to workspaces/items |
| `sql` | Azure SQL Database `user_impersonation` | SQL endpoint and data permissions |
| `graphql` | Power BI service `GraphQLApi.Execute.All` | API execution and source permissions |

The package qualifies permissions using each resource, including
`https://database.windows.net//user_impersonation` with SQL's required double
slash. See Microsoft's [Fabric data-plane guidance](https://learn.microsoft.com/en-us/fabric/workload-development-kit/fabric-data-plane)
and [Fabric REST scopes](https://learn.microsoft.com/en-us/rest/api/fabric/articles/scopes).
The client secret authenticates the Web app at the token endpoint. The resulting
data tokens represent the visitor, not a service principal or an OBO exchange.
See [Microsoft's authorization code flow](https://learn.microsoft.com/en-us/entra/identity-platform/v2-oauth2-auth-code-flow).

## Start with a configured SQL endpoint

This route needs only the SQL target and avoids workspace discovery permissions.
The complete app is installed under `examples/shiny-fabric`:

```{r, eval=FALSE}
example_dir <- system.file("examples", "shiny-fabric", package = "fabricQueryR")
shiny::runApp(example_dir, host = "127.0.0.1", port = 8100)
```

Set `ENTRA_TENANT_ID`, `ENTRA_CLIENT_ID`, `ENTRA_CLIENT_SECRET`,
`FABRIC_SQL_SERVER` and `FABRIC_SQL_DATABASE` first. The example uses DBI, odbc
and Microsoft ODBC Driver 18; it reads visible SQL table metadata with a bound
name-filter parameter. Its README describes the callback configuration.

The essential application wiring is:

```{r, eval=FALSE}
library(shiny)
library(fabricQueryR)

config <- fabric_shiny_config(
  tenant_id = Sys.getenv("ENTRA_TENANT_ID"),
  client_id = Sys.getenv("ENTRA_CLIENT_ID"),
  client_secret = Sys.getenv("ENTRA_CLIENT_SECRET"),
  redirect_uri = "http://localhost:8100/",
  services = "sql"
)
ui <- fabric_shiny_ui(
  fluidPage(actionButton("login", "Sign in"), tableOutput("tables")),
  "fabric", config
)
server <- function(input, output, session) {
  fabric <- fabric_shiny_server("fabric", config)
  observeEvent(input$login, fabric$login(), ignoreInit = TRUE)
  output$tables <- renderTable({
    req(fabric$ready("sql"))
    fabric_sql_query(
      Sys.getenv("FABRIC_SQL_SERVER"),
      "SELECT TABLE_SCHEMA, TABLE_NAME FROM INFORMATION_SCHEMA.TABLES",
      database = Sys.getenv("FABRIC_SQL_DATABASE"),
      token = fabric$token_provider(), numeric_policy = "driver"
    )
  })
}
shinyApp(ui, server, uiPattern = ".*")
```

Use `backend = "adbc"` with the corresponding driver installed when exact
numeric or Arrow results are needed. Query calls open and close their own
connections. Connections obtained from `fabric_sql_connect()` or `$sql_connect()`
must be disconnected explicitly; logout does not close an already authenticated
DBI connection. Avoid global per-user connection pools.

## Discover a Warehouse and reuse its methods

Configure `services = c("fabric", "sql")` to prepare both resource tokens from
one authorization. Inside the same server:

```{r, eval=FALSE}
warehouse <- reactive({
  req(fabric$ready("fabric"))
  fabric$item(
    workspace = Sys.getenv("FABRIC_WORKSPACE_ID"),
    item = Sys.getenv("FABRIC_WAREHOUSE_ID"),
    type = "Warehouse"
  )
})
output$orders <- renderTable({
  req(fabric$ready("sql"))
  warehouse()$sql_query(
    "SELECT TOP (100) order_id, region FROM dbo.orders WHERE region = ?",
    params = list(input$region), backend = "adbc"
  )
})
```

Supply an existing table and corresponding UI input for this example. GUIDs
avoid listing every workspace. `fabric$workspaces()` is available for an
application that wants discovery. Item methods retain the original credential.
When using standalone functions, always pass `token = fabric$token_provider()`,
including when an R6 item supplies the target. A missing `token` argument uses
the general package's AzureAuth behavior.

To avoid requiring unnecessary Fabric scopes, select the operations explicitly,
for example `scopes = list(fabric = "Item.Read.All")` for known item IDs. Scope
selection is fixed at registration/configuration time; `.default` consent,
automatic incremental consent, other service profiles and multiple tenants are
outside this first integration.

## GraphQL and an existing shinyOAuth app

A configured GraphQL URL can use only `services = "graphql"`. Call the existing
API with the session provider:

```{r, eval=FALSE}
result <- fabric_graphql_query(
  api = Sys.getenv("FABRIC_GRAPHQL_ENDPOINT"),
  query = configured_query_document,
  variables = list(region = input$region),
  token = fabric$token_provider()
)
```

When source permissions must follow the caller, configure the GraphQL source
connection appropriately and validate its SSO behavior. Saved source credentials
can establish a different authorization boundary. See Microsoft's
[GraphQL connection guidance](https://learn.microsoft.com/en-us/fabric/data-engineering/connect-apps-api-graphql).

Existing shinyOAuth applications can use the adapter without another login UI:

```{r, eval=FALSE}
# The existing client already declares the SQL resource and scope as target db.
provider <- reactive({
  connection <- req(auth$connection())
  fabric_shiny_token_provider(connection, services = "sql", targets = c(sql = "db"))
})
output$tables <- renderTable({
  fabric_sql_query(
    configured_server, "SELECT TABLE_NAME FROM INFORMATION_SCHEMA.TABLES",
    database = configured_database, token = provider(), numeric_policy = "driver"
  )
})
```

For several managed authorizations use `auth$connection(selected_id)` and keep
the manager's owner/activity requirements. The adapter captures that reference;
it never silently follows a later login. A Graph-only client cannot acquire SQL
merely by wrapping it. Configure the matching token targets before sign-in.

## Results, errors and lifetime

The server prefetches selected targets after login. `ready("sql")` checks
recorded scope coverage; it does not prove current network access or SQL grants.
Expired but refreshable tokens can recover on the next operation. Use
`status()` for local target state, and `prepare("sql")` to retry acquisition.
Tokens are not included in status. An optional target failure does not discard
another service's usable authorization.

| Outcome | Application action |
| --- | --- |
| Signed out | Show a button calling `login()` |
| `interaction_required` | Offer `reauthorize()` explicitly |
| `insufficient_scope` | Check declared permissions and consent; avoid refresh loops |
| `refresh_pending` | Offer a retry; synchronous calls cannot wait on a promise |
| Fabric/SQL permission denial | Explain data access is denied; repeated login is not a remedy |
| Unavailable/ended reference | Discard the object and obtain a new session provider |

Standalone calls preserve `shinyOAuth_access_error` and its `context$reason`.
SQL can wrap acquisition failures in a `fabric_sql_authentication_error` parent
chain. Do not show raw provider diagnostics or tokens in the browser. Successful
silent refresh does not change `generation()` or rerun generation-dependent
queries. Explicit status reads can update on refresh.

HTTP calls retain fabricQueryR's retry behavior: one 401 can cause a forced
token acquisition and request retry even with `idempotent = FALSE`. That is
separate from transient network/service retries. `prepare()` and
`reauthorize()` themselves do not replay data operations.

Bind event-loaded results to their authorization before rendering:

```{r, eval=FALSE}
rows <- reactiveVal(NULL)
observeEvent(fabric$generation(), rows(NULL), ignoreNULL = FALSE, priority = 50)
observeEvent(input$load, {
  req(fabric$ready("sql"))
  generation <- fabric$generation()
  data <- warehouse()$sql_query(configured_read_query)
  if (identical(generation, fabric$generation())) {
    rows(list(generation = generation, data = data))
  }
}, ignoreInit = TRUE)
output$rows <- renderTable({
  req(fabric$ready("sql"))
  value <- req(rows())
  req(identical(value$generation, fabric$generation()))
  value$data
})
```

Avoid `cancelOutput = TRUE`, which keeps old output visible. Keep cached data
session-local, and include authorization generation plus query parameters in
cache keys. Do not send providers or R6 items to futures or other processes.
This release does not provide a nonblocking task/credential-transfer interface.

Small writes can use a short DBI connection with bound parameters and
`on.exit(DBI::dbDisconnect(con))`, inside an explicit Save event. They require a
writable endpoint and the user's write grants. Do not replay a write after
reauthorization or an ambiguous timeout. Reader/writer acceptance is pending;
the supplied runnable example performs reads only.

## Hosting and remaining acceptance

Use one R process, with the public HTTPS callback/base path registered exactly.
Keep callback HTTP requests and the owning WebSocket session routed correctly.
The UI wrapper reuses shinyOAuth's state/PKCE/nonce routing; `uiPattern = ".*"`
allows callback subpaths to reach it. Configure reverse proxies and cookies
according to the shinyOAuth deployment guidance for your chosen public URL.
The wrapper does not install a proxy route. Test the deployed callback URL.

The server defaults to an eight-hour local authorization age; set
`reauth_after_seconds` for your app. Closing the Shiny session ends its references.
Local logout does not promise global Microsoft sign-out. Cross-process storage,
retained authorizations and asynchronous worker renewal require separate designs.

Standard Microsoft hosts are allowed per service. For a trusted gateway, declare
its exact hostname through `endpoint_hosts`. Never accept that list from a user.
The provider's policy survives R6 credentials and SQL retries. HTTP redirects
are rejected; continuation/operation endpoints are checked before acquisition.
These helpers are not an isolation mechanism for untrusted R code.

Before treating a deployment as accepted, perform the delegated browser checks
recorded in `tools/design/shiny-fabric-acceptance.md` in the source repository:
two users with different data permissions, logout/account replacement, token
refresh, reader/writer separation, denied consent, callback subpaths and GraphQL
source authorization. None of those live checks is implied by an offline pass.
