---
title: "Shiny apps using each visitor's Fabric access"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Shiny apps using each visitor's Fabric access}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

fabricQueryR lets you build Shiny apps around data in Microsoft Fabric.
You can use a regular Fabric connection with a shared account or the app's own
credentials; queries then use that connection's permissions.

This vignette focuses on delegated access: each visitor signs in, and the
Shiny helpers shown here use that user's authorization for Fabric queries.
The advantage is that the permissions you already manage in Fabric determine
which data each user can access through the app. You can reuse those access
rules without maintaining a second set in Shiny.

You can work with Warehouse and Lakehouse tables using SQL, semantic models
using DAX, Eventhouse data using KQL, files and Delta tables in OneLake, or an
API for GraphQL.

This guide starts with SQL: connect to a Warehouse or Lakehouse, run a query,
and display the result in Shiny. It then shows how to build a dashboard from a
semantic model using DAX, reusing the measures and relationships used by Power
BI. The complete app includes both approaches and the other data sources, so
you can see how each fits into a Shiny app.

Queries run in background R processes using Shiny's `ExtendedTask`, keeping
the app responsive while data loads. The shinyOAuth package handles Microsoft
sign-in.

## Setup

### Install the packages

This integration currently requires the development versions of fabricQueryR
and shinyOAuth. If you have both source checkouts, install shinyOAuth first:

```{r, eval=FALSE}
devtools::install("../shinyoauth")
devtools::install(".")
install.packages(c("shiny", "bslib", "mirai", "promises", "DBI", "odbc"))
```

Use shiny 1.8.1 or later, bslib 0.7.0 or later, and promises 1.3.0 or later.
The app uses mirai when installed. Without mirai, install future to use the
`future_promise()` fallback. For SQL, install Microsoft ODBC Driver 18 for SQL
Server separately on the computer running the app and its workers. DAX, KQL
and GraphQL use HTTP APIs, so they do not need a database driver. OneLake's
additional dependencies are described under Other data sources below.

### Arrange Microsoft sign-in

Your app needs a Web application registration in Microsoft Entra, the service
that manages your organization's Microsoft sign-in. Your Fabric or Microsoft
365 administrator can help set this up.

For SQL, add the Azure SQL Database delegated permission
`user_impersonation`. A delegated permission allows the app to make requests
using the signed-in person's access. Users also need permission to connect
to the database and read the requested data.

The registration needs a redirect URI: the address where Microsoft sends
the visitor after sign-in. For local development, use `http://localhost:8100/`.
The address in the registration and the app must match. A deployed app uses
its public HTTPS address instead.

You will need the tenant ID, which identifies your organization's directory;
the client ID, which identifies your app; and a client secret, which the app
uses during sign-in. Complete any administrator consent your organization
requires before running the app.

### Save the connection settings

The app reads settings with `Sys.getenv()`. For local development, put these
in your user `.Renviron` file and restart R:

```text
ENTRA_TENANT_ID=your-directory-id
ENTRA_CLIENT_ID=your-application-id
ENTRA_CLIENT_SECRET=your-client-secret
FABRIC_WAREHOUSE_SQL_SERVER=your-warehouse-sql-hostname
FABRIC_WAREHOUSE_SQL_DATABASE=your-warehouse-database
```

Copy the server and database names from your Warehouse's SQL connection
details. The Lakehouse section below gives the equivalent settings for its SQL
analytics endpoint. Keep the client secret out of app code and version control;
deployed apps can use the hosting service's environment-variable or secret
settings.

## A complete example app

Save the following code as `app.R` in a new directory. Keep the entries you
want in `sources`; to start with just a Warehouse, keep
`"Warehouse (SQL)" = "warehouse"`. The sign-in configuration automatically
selects the services needed by those entries. If you keep other sources, add
their settings and registration permissions as described below.

The SQL queries use an example `dbo.Sales` table; replace it with your own table.
The other branches also use example table, column and measure names, described
in their sections below.

```{r fabric-app, eval=FALSE}
library(shiny)
library(fabricQueryR)

use_mirai <- requireNamespace("mirai", quietly = TRUE)
if (use_mirai) {
  mirai::daemons(2)
  onStop(\() mirai::daemons(0), session = NULL)
} else {
  future::plan(future::multisession, workers = 2)
  onStop(\() future::plan(future::sequential), session = NULL)
}

# Keep the sources your app needs.
sources <- c(
  "Warehouse (SQL)" = "warehouse",
  "Lakehouse (SQL)" = "lakehouse",
  "Semantic model (DAX)" = "dax",
  "Eventhouse (KQL)" = "kql",
  "OneLake file" = "files",
  "OneLake Delta table" = "delta",
  "GraphQL API" = "graphql"
)
service_for <- c(
  warehouse = "sql",
  lakehouse = "sql",
  dax = "dax",
  kql = "kql",
  files = "onelake",
  delta = "onelake",
  graphql = "graphql"
)
config <- fabric_shiny_config(
  tenant_id = Sys.getenv("ENTRA_TENANT_ID"),
  client_id = Sys.getenv("ENTRA_CLIENT_ID"),
  client_secret = Sys.getenv("ENTRA_CLIENT_SECRET"),
  redirect_uri = Sys.getenv("ENTRA_REDIRECT_URI", "http://localhost:8100/"),
  services = unique(unname(service_for[sources]))
)

# Replace the example table, column and measure names with your own.
read_data <- function(source, token, year) {
  switch(
    source,
    warehouse = fabric_sql_query(
      server = Sys.getenv("FABRIC_WAREHOUSE_SQL_SERVER"),
      database = Sys.getenv("FABRIC_WAREHOUSE_SQL_DATABASE"),
      sql = "SELECT TOP (100) * FROM dbo.Sales",
      token = token,
      numeric_policy = "driver",
      verbose = FALSE
    ),
    lakehouse = fabric_sql_query(
      server = Sys.getenv("FABRIC_LAKEHOUSE_SQL_SERVER"),
      database = Sys.getenv("FABRIC_LAKEHOUSE_SQL_DATABASE"),
      sql = "SELECT TOP (100) * FROM dbo.Sales",
      token = token,
      numeric_policy = "driver",
      verbose = FALSE
    ),
    dax = {
      data <- fabric_pbi_dax_query(
        workspace_id = Sys.getenv("FABRIC_WORKSPACE_ID"),
        dataset_id = Sys.getenv("FABRIC_SEMANTIC_MODEL_ID"),
        dax = sprintf(
          r"(
          EVALUATE SUMMARIZECOLUMNS(
            'Product'[Category], TREATAS({%d}, 'Date'[Year]),
            "Sales", [Total Sales]
          )
        )",
          as.integer(year)
        ),
        token = token
      )
      if (ncol(data) == 2L) {
        names(data) <- c("Category", "Sales")
      }
      data
    },
    kql = fabric_kql_query(
      cluster = Sys.getenv("FABRIC_KQL_ENDPOINT"),
      database = Sys.getenv("FABRIC_KQL_DATABASE"),
      query = "Sales | take 100",
      token = token
    ),
    files = fabric_onelake_read_file(
      workspace = Sys.getenv("FABRIC_WORKSPACE_ID"),
      item = Sys.getenv("FABRIC_LAKEHOUSE_ID"),
      path = Sys.getenv("FABRIC_ONELAKE_FILE", "Files/sales.parquet"),
      token = token
    ),
    delta = fabric_onelake_read_delta_table(
      table_path = Sys.getenv("FABRIC_DELTA_TABLE", "Sales"),
      schema = Sys.getenv("FABRIC_DELTA_SCHEMA", "dbo"),
      workspace_name = Sys.getenv("FABRIC_WORKSPACE_ID"),
      lakehouse_name = Sys.getenv("FABRIC_LAKEHOUSE_ID"),
      limit = 100,
      token = token,
      verbose = FALSE
    ),
    graphql = {
      result <- fabric_graphql_query(
        api = Sys.getenv("FABRIC_GRAPHQL_ENDPOINT"),
        query = "query { sales(first: 100) { items { Category Amount } } }",
        error_policy = "error",
        token = token
      )
      do.call(rbind, lapply(result$data$sales$items, as.data.frame))
    }
  )
}

ui <- fabric_shiny_ui(
  fluidPage(
    titlePanel("Explore Fabric data"),
    actionButton("login", "Sign in"),
    actionButton("logout", "Sign out"),
    selectInput("source", "Data source", sources),
    conditionalPanel(
      "input.source === 'dax'",
      numericInput("year", "Year", 2026, min = 1900, max = 2100, step = 1)
    ),
    bslib::input_task_button("load", "Load data"),
    conditionalPanel("input.source === 'dax'", plotOutput("sales")),
    tableOutput("tables")
  ),
  "fabric",
  config
)

server <- function(input, output, session) {
  fabric <- fabric_shiny_server("fabric", config)
  query <- ExtendedTask$new(function(token, generation, source, year) {
    promises::then(token, function(token) {
      data <- if (use_mirai) {
        mirai::mirai(
          {
            library(fabricQueryR)
            read_data(source, token, year)
          },
          read_data = read_data,
          source = source,
          token = token,
          year = year
        )
      } else {
        promises::future_promise(read_data(source, token, year))
      }
      promises::then(data, function(data) {
        list(generation = generation, source = source, year = year, data = data)
      })
    })
  }) |>
    bslib::bind_task_button("load")

  observeEvent(input$login, fabric$login(), ignoreInit = TRUE)
  observeEvent(input$logout, fabric$logout(), ignoreInit = TRUE)
  observeEvent(
    input$load,
    {
      service <- service_for[[req(input$source)]]
      req(fabric$ready(service))
      query$invoke(
        fabric$access_token(service, async = TRUE),
        fabric$generation(),
        input$source,
        input$year
      )
    },
    ignoreInit = TRUE
  )

  result <- reactive({
    req(fabric$ready(service_for[[req(input$source)]]))
    value <- query$result()
    req(identical(value$generation, fabric$generation()))
    req(identical(value$source, input$source))
    if (input$source == "dax") {
      req(identical(value$year, input$year))
    }
    value$data
  })
  output$tables <- renderTable(head(result(), 100))
  output$sales <- renderPlot({
    req(input$source == "dax")
    data <- result()
    req(nrow(data) > 0)
    barplot(as.numeric(data$Sales), names.arg = data$Category, ylab = "Sales")
  })
}

shinyApp(
  ui,
  server,
  uiPattern = ".*",
  options = list(host = "127.0.0.1", port = 8100)
)
```

Run the directory with `shiny::runApp("path/to/your/app")`, then open
`http://localhost:8100/`. Click Sign in, leave Warehouse (SQL) selected, and
click Load data. The app displays up to 100 rows from your table. You can then
adapt the query or choose another configured source.

### SQL: Warehouse and Lakehouse

The two SQL entries use `fabric_sql_query()` with different endpoints. Both
return an R table for `renderTable()`. The Warehouse uses the settings from
Setup. To use the Lakehouse entry, copy the server and database names from its
SQL analytics endpoint's connection details:

```text
FABRIC_LAKEHOUSE_SQL_SERVER=your-lakehouse-sql-hostname
FABRIC_LAKEHOUSE_SQL_DATABASE=your-lakehouse-database
```

Replace `dbo.Sales` with your table. For filters taken from Shiny inputs, use
SQL parameters: for example, `sql = "SELECT * FROM dbo.Sales WHERE Region = ?"`
with `params = list(region)`. Pass `region` to the task as an ordinary value.

Both SQL branches use `numeric_policy = "driver"`, which keeps the ODBC
driver's numeric representation. For exact numeric or Arrow results, see
[SQL queries](reading-data.html) and the ADBC backend.

### DAX: a semantic-model dashboard

To use the semantic-model entry, add these settings to `.Renviron` and restart R:

```text
FABRIC_WORKSPACE_ID=your-workspace-id
FABRIC_SEMANTIC_MODEL_ID=your-semantic-model-id
```

Use the semantic model's ID, rather than a report's ID. Add the Power BI Service
delegated permission `Dataset.Read.All` to the registration. Each user needs
Read and Build permission on the model, and the tenant must allow the Dataset
Execute Queries REST API. See Microsoft's
[DAX query requirements](https://learn.microsoft.com/en-us/rest/api/power-bi/datasets/execute-queries).

The example assumes `'Product'[Category]`, `'Date'[Year]` and a `[Total Sales]`
measure. Replace these with names from your model. Select Semantic model (DAX)
in the app, choose a year present in the model, and click Load data to display
the sales chart. Change the year and click Load data again to update it.

`fabric_pbi_dax_query()` sends DAX to an existing semantic model and returns
an R table. In this example, `SUMMARIZECOLUMNS()` groups by product category,
`TREATAS()` applies the selected year, and `[Total Sales]` is evaluated by the
model. You can reuse business measures already defined there while designing
the dashboard in Shiny.

The app renames the two returned columns to `Category` and `Sales` for use
with `renderTable()` and `renderPlot()`. Replace the base R chart with ggplot2
or another Shiny output as needed. For a larger dashboard, use separate tasks
for views that should load independently, or return one dataset that several
outputs can share.

Requests run as the signed-in user; model permissions and any row-level
security applicable to that user govern the result. The model remains the
source of measures and relationships, while Shiny provides the dashboard UI.

### How the background query works

The app uses three fabricQueryR functions to connect sign-in with your Shiny UI:

- `fabric_shiny_config()` holds the sign-in settings and selects the services.
  Create it once, outside `server()`.
- `fabric_shiny_ui()` wraps your usual UI so the app can complete sign-in.
- `fabric_shiny_server()` creates helpers for the current visitor. Call it
  inside `server()` so each visitor has their own connection.

The Load data button starts an `ExtendedTask`. First,
`fabric$access_token(service, async = TRUE)` acquires the visitor's token for
the selected service. Then mirai runs `read_data()` in a background R process,
or `future_promise()` does so if mirai is not installed. `query$result()` makes
the result available to the Shiny outputs when it is ready. The task button
stays busy until that work completes.

Read reactive inputs when calling `$invoke()`, before the worker starts.
The example passes the source and year as ordinary values. It also passes
`fabric$generation()`, which identifies the current sign-in. The result checks
prevent data from an earlier sign-in or a different source or year being
displayed. Signing out clears the outputs; an already running query can still
finish. Signing in again requires a new click on Load data.

See Shiny's [nonblocking operations guide](https://shiny.posit.co/r/articles/improve/nonblocking/)
for more about `ExtendedTask`, and the
[mirai Shiny guide](https://mirai.r-lib.org/articles/mirai-promises.html)
for its background workers. The mirai call explicitly supplies the reader
function and query inputs, and loads fabricQueryR in the worker.

## Other data sources

Each branch of `read_data()` uses a regular fabricQueryR function with the
current user's token. You can keep one branch, combine several sources in a
dashboard, or replace the example queries. The displayed table shows at most
100 rows; the OneLake file branch reads the whole file before displaying it.

### KQL: Eventhouse

Set the query URI and database from your KQL database's connection details:

```text
FABRIC_KQL_ENDPOINT=https://your-cluster.kusto.fabric.microsoft.com
FABRIC_KQL_DATABASE=your-kql-database
```

The app runs `Sales | take 100` with `fabric_kql_query()`. Replace `Sales` with
your table or use your own KQL query. The registration uses Azure Data Explorer's
delegated `user_impersonation` permission, and each user needs access to the
database. See [KQL queries](reading-data.html) for query parameters and results.

### OneLake files and Delta tables

Set the workspace and Lakehouse IDs, together with the file or table to read:

```text
FABRIC_WORKSPACE_ID=your-workspace-id
FABRIC_LAKEHOUSE_ID=your-lakehouse-id
FABRIC_ONELAKE_FILE=Files/sales.parquet
FABRIC_DELTA_TABLE=Sales
FABRIC_DELTA_SCHEMA=dbo
```

`fabric_onelake_read_file()` reads a Parquet, CSV or Arrow file into R.
Install the arrow package for this branch. `fabric_onelake_read_delta_table()`
reads a Delta table directly from OneLake, with `limit = 100` in this example.
Set the schema separately; use `FABRIC_DELTA_SCHEMA=` for a Lakehouse without
schemas. The Delta reader needs reticulate and a compatible Python environment.
`fabric_delta_config(initialize = TRUE)` prepares the runtime; see
[OneLake and Lakehouse](onelake-and-lakehouse.html) for details.

These branches use Azure Storage's delegated `user_impersonation` permission
and the user's OneLake access. SQL and direct Delta reads offer different ways
to work with Lakehouse tables; choose the interface that suits your app.

### GraphQL

Set the URL of an existing Fabric API for GraphQL:

```text
FABRIC_GRAPHQL_ENDPOINT=https://your-api.graphql.fabric.microsoft.com/graphql
```

The example calls `fabric_graphql_query()` and turns the returned `sales`
items into an R data frame. Replace the query and `result$data$sales$items`
with the names in your API's schema. `error_policy = "error"` reports a failed
GraphQL query through the task's output. Add the Power BI Service delegated
permission `GraphQLApi.Execute.All` to the registration.

The API's source connection determines access to the underlying data. To use
each visitor's source permissions, configure its single sign-on settings
accordingly. See Microsoft's
[GraphQL connection guidance](https://learn.microsoft.com/en-us/fabric/data-engineering/connect-apps-api-graphql)
and [GraphQL queries](graphql-schema-and-rows.html).

## Advanced configuration

### Discovery and service permissions

The example uses known endpoints and IDs. Add `"fabric"` to `services` if you
also want to discover workspaces and items. Its default permissions are
`Workspace.Read.All` and `Item.Read.All`; you can choose a narrower set with
`scopes = list(fabric = "Item.Read.All")` when using known item IDs.

A task that discovers an item and queries it needs a token for each service.
For example, acquire both tokens in the Shiny process:

```{r, eval=FALSE}
tokens <- promises::promise_all(
  fabric = fabric$access_token("fabric", async = TRUE),
  sql = fabric$access_token("sql", async = TRUE)
)
```

Resolve that promise before starting the background query, as the main app
does for one token. With mirai, pass the resolved list as `tokens = tokens`.
Inside the worker, create the item and query it:

```{r, eval=FALSE}
warehouse <- fabric_item(
  workspace = Sys.getenv("FABRIC_WORKSPACE_ID"),
  item = Sys.getenv("FABRIC_WAREHOUSE_ID"),
  type = "Warehouse", token = tokens$fabric
)
data <- warehouse$sql_query(
  "SELECT TOP (100) * FROM dbo.Sales",
  token = tokens$sql, numeric_policy = "driver", verbose = FALSE
)
```

The package fills in full scope URLs, including SQL's
`https://database.windows.net//user_impersonation`. DAX and GraphQL share a
Power BI token target; when both services are selected, their permissions are
combined automatically. See [Fabric REST scopes](https://learn.microsoft.com/en-us/rest/api/fabric/articles/scopes)
when choosing discovery permissions.

Sign-in also requests `openid`, `profile` and `offline_access` to identify the
visitor and renew credentials. Identity comes from a validated ID token;
Graph UserInfo and `User.Read` are not needed just to display the visitor's
name. The client secret authenticates the app during sign-in, while data
requests use the visitor's delegated access.

### Using an existing shinyOAuth app

If your app already handles sign-in with shinyOAuth, its connection can supply
a token to the same `ExtendedTask`. Declare the required token targets on the
shinyOAuth client before sign-in. For a DAX target named `power_bi`, invoke
the task with:

```{r, eval=FALSE}
observeEvent(input$load, {
  connection <- req(auth$connection())
  query$invoke(
    connection$access_token(
      target = "power_bi",
      required_scopes = "https://analysis.windows.net/powerbi/api/Dataset.Read.All",
      min_valid_for = 60, async = TRUE
    ),
    connection$id, "dax", input$year
  )
}, ignoreInit = TRUE)
```

Adapt the result check to compare `value$generation` with the current
`auth$connection()$id`, and keep the year check. For several managed
authorizations, select one with `auth$connection(selected_id)` and follow
shinyOAuth's ownership and activity requirements. For synchronous calls in
the Shiny process, `fabric_shiny_token_provider()` adapts an existing
connection to the usual `token` argument.

### Diagnosing connection problems

`fabric$ready("dax")` checks the permissions recorded for the connection. A
query can still fail because of service settings, network problems or data
permissions. Use `fabric$status()` to inspect connection state and
`fabric$prepare("dax")` to retry preparing DAX access. A problem with one
service can leave the others available. Use `fabric$reauthorize()` to start
sign-in again.

For application error handling, `shinyOAuth_access_error` provides a
`context$reason` field. `interaction_required` calls for explicit sign-in
again; `insufficient_scope` calls for checking permissions and consent.
`refresh_pending` means a synchronous call cannot yet get the renewed
credential. SQL may wrap these errors in the parent of a
`fabric_sql_authentication_error`.

Routine token refresh does not change `generation()` or rerun queries.
A token passed to a worker is fixed for that invocation; it does not renew
there. If it expires, start a new task to acquire a fresh token.
Synchronous calls using a token provider retain fabricQueryR's retry behavior:
one HTTP 401 response can trigger token renewal and a request retry.
`prepare()` and `reauthorize()` themselves do not replay data operations.

### Connections and deployment

`fabric_sql_query()` opens and closes its own connection. If you use
`fabric_sql_connect()` or `$sql_connect()` directly, close it with
`DBI::dbDisconnect()`, usually through `on.exit()` in a helper function.
Signing out does not close an already open DBI connection.

Set up the worker pool once when the app starts. The example uses two
background R processes: `mirai::daemons(2)` when mirai is installed, or
`future::plan(future::multisession, workers = 2)` otherwise. Adjust that count
for your host. The workers serve all visitors, and `onStop()` shuts them down
when the app stops. Return ordinary R data from a worker; keep live database
connections and Arrow streams within the process that created them.

Use an exact registered HTTPS callback URL for deployment.
The `uiPattern = ".*"` setting lets callback subpaths reach the app. Check your
hosting service's routing, reverse-proxy and cookie settings using shinyOAuth's
deployment guidance. The default local sign-in lifetime is eight hours; adjust
`reauth_after_seconds` for your app. App logout does not sign the visitor out
of Microsoft everywhere.

Standard Microsoft service hosts are allowed. A trusted gateway needs its
exact hostname in `endpoint_hosts`, configured by the app author.
