---
title: "Shiny apps using each visitor's Fabric access"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Shiny apps using each visitor's Fabric access}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

fabricQueryR lets Shiny apps query Microsoft Fabric using each signed-in user's
permissions. Visitors sign in with a Microsoft account, and requests to Fabric
use that account's access to the data.

The example below runs queries in the background using Shiny's `ExtendedTask`,
so the app stays responsive while Fabric returns data. You can display the
results with Shiny functions such as `renderTable()`. The shinyOAuth package
handles the Microsoft sign-in process.

This vignette starts with an app that lists the tables a user can see in a
Fabric SQL database. It then shows how to work with a Warehouse, load data on
demand, and adapt an app that already uses shinyOAuth.

## Setup

### Install the packages

This integration currently requires the development versions of fabricQueryR
and shinyOAuth. If you have both source checkouts, install shinyOAuth first:

```{r, eval=FALSE}
devtools::install("../shinyoauth")
devtools::install(".")
```

For the SQL example, you also need shiny (1.8.1 or later), bslib (0.7.0 or
later), future, promises, DBI, odbc, and Microsoft ODBC Driver 18 for SQL Server.
The driver is installed separately from the R packages, on the computer that
runs the app and its background workers.

### Arrange Microsoft sign-in

Your app needs an application registration in Microsoft Entra, the service that
manages your organization's Microsoft sign-in. Your Fabric or Microsoft 365
administrator can help set this up.

For the SQL example, ask for a Web application registration with the Azure
SQL Database delegated permission `user_impersonation`. A delegated permission
allows the app to make requests using the signed-in person's access. Users still
need their own permission to connect to the database and read its data.

The registration needs a redirect URI: the address where Microsoft sends
the visitor after sign-in. For local development, use `http://localhost:8100/`.
The address in the registration and the app must match. A deployed app uses its
public HTTPS address instead.

You will need three values from the registration: the tenant ID, which
identifies your organization's directory; the client ID, which identifies
your app; and a client secret, which the app uses during sign-in. Complete
any administrator consent your organization requires before testing the app.

### Save the connection settings

The examples read settings with `Sys.getenv()`. For local development, you can
put them in your user `.Renviron` file and restart R:

```text
ENTRA_TENANT_ID=your-directory-id
ENTRA_CLIENT_ID=your-application-id
ENTRA_CLIENT_SECRET=your-client-secret
FABRIC_SQL_SERVER=your-sql-endpoint-hostname
FABRIC_SQL_DATABASE=your-database-name
```

Copy the SQL server and database name from the connection details for your
Fabric Warehouse, SQL Database or Lakehouse SQL analytics endpoint. Keep the
client secret out of your app code and version control. For a deployed app,
use the hosting service's environment-variable or secret settings.

## A complete example app

After saving the settings above, save the following code as `app.R` in a new
directory. This is the complete app: it signs visitors in, loads their table
list in the background, and lets them sign out.

```{r fabric-app, eval=FALSE}
library(shiny)
library(fabricQueryR)

future::plan(future::multisession, workers = 2)

config <- fabric_shiny_config(
  tenant_id = Sys.getenv("ENTRA_TENANT_ID"),
  client_id = Sys.getenv("ENTRA_CLIENT_ID"),
  client_secret = Sys.getenv("ENTRA_CLIENT_SECRET"),
  redirect_uri = Sys.getenv("ENTRA_REDIRECT_URI", "http://localhost:8100/"),
  services = "sql"
)
sql_server <- Sys.getenv("FABRIC_SQL_SERVER")
sql_database <- Sys.getenv("FABRIC_SQL_DATABASE")

ui <- fabric_shiny_ui(
  fluidPage(
    actionButton("login", "Sign in"),
    actionButton("logout", "Sign out"),
    textInput("table_name", "Table name contains", ""),
    bslib::input_task_button("load", "Show tables"),
    tableOutput("tables")
  ),
  "fabric",
  config
)

server <- function(input, output, session) {
  fabric <- fabric_shiny_server("fabric", config)
  query <- ExtendedTask$new(function(token, generation, table_name) {
    promises::then(token, function(token) {
      promises::future_promise({
        data <- fabric_sql_query(
          sql_server,
          paste(
            "SELECT TOP (100) TABLE_SCHEMA, TABLE_NAME",
            "FROM INFORMATION_SCHEMA.TABLES WHERE TABLE_NAME LIKE ?",
            "ORDER BY TABLE_SCHEMA, TABLE_NAME"
          ),
          database = sql_database,
          params = list(paste0("%", table_name, "%")),
          token = token,
          numeric_policy = "driver",
          verbose = FALSE
        )
        list(generation = generation, data = data)
      })
    })
  }) |>
    bslib::bind_task_button("load")

  observeEvent(input$login, fabric$login(), ignoreInit = TRUE)
  observeEvent(input$logout, fabric$logout(), ignoreInit = TRUE)
  observeEvent(
    input$load,
    {
      req(fabric$ready("sql"))
      query$invoke(
        fabric$access_token("sql", async = TRUE),
        fabric$generation(),
        input$table_name
      )
    },
    ignoreInit = TRUE
  )

  output$tables <- renderTable({
    req(fabric$ready("sql"))
    result <- query$result()
    req(identical(result$generation, fabric$generation()))
    result$data
  })
}

shinyApp(
  ui,
  server,
  uiPattern = ".*",
  options = list(host = "127.0.0.1", port = 8100)
)
```

Run the directory with `shiny::runApp("path/to/your/app")`, then open
`http://localhost:8100/` and click Sign in, then Show tables. You can enter part
of a table name to filter the result. While the query runs, the button shows its
progress and the rest of the app stays responsive. An empty result can be valid
if the account cannot see any matching tables. Click Sign out to end the app's
sign-in and clear the table display.

The app uses three fabricQueryR functions:

- `fabric_shiny_config()` holds the sign-in settings and selects the services
  the app will use. Create it once, outside `server()`.
- `fabric_shiny_ui()` wraps your usual UI so the app can complete sign-in.
- `fabric_shiny_server()` creates helpers for the current visitor. Call it
  inside `server()` so each visitor has their own connection.

The Sign in button calls `fabric$login()`. The Show tables button starts the
`ExtendedTask` with `query$invoke()`. `fabric$access_token("sql", async = TRUE)`
acquires the visitor's SQL token without waiting in the Shiny event handler.
Once it is available, `future_promise()` runs the query in a background R
process. Fabric checks the visitor's permission to access the requested data.

`query$result()` makes the result available to `renderTable()` when it is ready.
The task button is bound to the task, so it stays busy until the query completes.
See Shiny's [nonblocking operations guide](https://shiny.posit.co/r/articles/improve/nonblocking/)
for more about `ExtendedTask`.

To adapt this example, replace the SQL with your own read query and add the
inputs and outputs your app needs. Use `params` for values supplied by
inputs, as in the Warehouse example below.

## Discover a Warehouse and reuse its methods

If you know a Warehouse's workspace ID and item ID, you can look it up in the
background task and use its usual fabricQueryR methods. Change the configuration
to `services = c("fabric", "sql")`: `"fabric"` lets the app look up the item,
and `"sql"` lets it query the data.

The example below assumes an existing `dbo.orders` table, a Shiny input named
`region`, and the environment variables `FABRIC_WORKSPACE_ID` and
`FABRIC_WAREHOUSE_ID`. Replace the task definition and Show tables observer in
the app above with:

```{r, eval=FALSE}
query <- ExtendedTask$new(function(tokens, generation, region) {
  promises::then(tokens, function(tokens) {
    promises::future_promise({
      warehouse <- fabric_item(
        workspace = Sys.getenv("FABRIC_WORKSPACE_ID"),
        item = Sys.getenv("FABRIC_WAREHOUSE_ID"),
        type = "Warehouse", token = tokens$fabric
      )
      data <- warehouse$sql_query(
        "SELECT TOP (100) order_id, region FROM dbo.orders WHERE region = ?",
        params = list(region), token = tokens$sql,
        numeric_policy = "driver", verbose = FALSE
      )
      list(generation = generation, data = data)
    })
  })
}) |> bslib::bind_task_button("load")

observeEvent(input$load, {
  req(fabric$ready())
  query$invoke(
    promises::promise_all(
      fabric = fabric$access_token("fabric", async = TRUE),
      sql = fabric$access_token("sql", async = TRUE)
    ),
    fabric$generation(), input$region
  )
}, ignoreInit = TRUE)
```

The task receives a token for each service. It creates the Warehouse object in
the worker, then passes the SQL token to `$sql_query()`. The table output from
the first example also works here. To use the ADBC backend instead, install its
driver and add `backend = "adbc"` to the query.

## Use GraphQL

An existing Fabric API for GraphQL is another way to query data, without a SQL
driver. Configure `services = "graphql"`, change the readiness checks to
`fabric$ready("graphql")`, and acquire `fabric$access_token("graphql", async = TRUE)`
in the Show tables observer. Inside the task's `future_promise()` block, replace
the SQL call with a GraphQL call:

```{r, eval=FALSE}
data <- fabric_graphql_query(
  api = Sys.getenv("FABRIC_GRAPHQL_ENDPOINT"),
  query = configured_query_document,
  variables = list(name = table_name),
  token = token
)
```

Replace `configured_query_document` and its variables with a query for your API's
schema, and adapt the output to the returned data. If the
underlying data source should use each visitor's permissions, check that the
API's source connection uses the appropriate single sign-on settings. A source
connection using saved credentials can give different access. See Microsoft's
[GraphQL connection guidance](https://learn.microsoft.com/en-us/fabric/data-engineering/connect-apps-api-graphql).

## Task inputs and sign-out

Use `fabric$logout()` to sign out, as in the example above, or
`fabric$reauthorize()` to start sign-in again.

An `ExtendedTask` uses the values passed to `$invoke()`. Read inputs such as
`input$table_name` there, before the worker starts. Editing an input while a
query runs affects the next invocation.

`fabric$generation()` identifies the current sign-in. The example passes it to
the task and checks it before displaying the result. A query already running
can finish after sign-out, but its result will no longer be displayed. Signing
in again does not rerun the query; click Show tables to load a new result.

## Advanced configuration

The following details are useful when adapting an existing authentication setup,
troubleshooting access, or preparing an app for deployment.

### Choosing permissions

`services` selects the kinds of Fabric connection your app needs. The SQL
example uses only `"sql"`; adding `"fabric"` requests permissions to look up
workspaces and items. Its defaults are `Workspace.Read.All` and `Item.Read.All`.
You can request a narrower set with, for example,
`scopes = list(fabric = "Item.Read.All")` when using known item IDs. GraphQL
uses the Power BI service permission `GraphQLApi.Execute.All`.

The package fills in the full scope URLs, including SQL's
`https://database.windows.net//user_impersonation`. You normally do not need to
construct these yourself. See [Fabric data-plane authentication](https://learn.microsoft.com/en-us/fabric/workload-development-kit/fabric-data-plane)
and [Fabric REST scopes](https://learn.microsoft.com/en-us/rest/api/fabric/articles/scopes)
when choosing permissions for a registration.

Sign-in also requests `openid`, `profile` and `offline_access` to identify the
visitor and renew credentials. Identity comes from a validated ID token; Graph
UserInfo and `User.Read` are not needed just to display the visitor's name.
The client secret authenticates the app during sign-in, while data requests use
the visitor's delegated access. This follows Microsoft's
[authorization code flow](https://learn.microsoft.com/en-us/entra/identity-platform/v2-oauth2-auth-code-flow).

### Using an existing shinyOAuth app

If your app already handles sign-in with shinyOAuth, use
its existing connection to supply the token to the same `ExtendedTask`. The
shinyOAuth client must declare a token target for the service you want to use.
For a SQL target named `db`, the invocation and result display become:

```{r, eval=FALSE}
observeEvent(input$load, {
  connection <- req(auth$connection())
  query$invoke(
    connection$access_token(
      target = "db",
      required_scopes = "https://database.windows.net//user_impersonation",
      min_valid_for = 60, async = TRUE
    ),
    connection$id, input$table_name
  )
}, ignoreInit = TRUE)
output$tables <- renderTable({
  connection <- req(auth$connection())
  result <- query$result()
  req(identical(result$generation, connection$id))
  result$data
})
```

For several managed authorizations, select one with
`auth$connection(selected_id)` and follow shinyOAuth's ownership and activity
requirements. Configure the needed targets before sign-in. For synchronous
calls in the Shiny process, `fabric_shiny_token_provider()` also adapts an
existing connection to the usual `token` argument.

### Diagnosing connection problems

`fabric$ready("sql")` checks the permissions recorded for the connection. The
query itself can still fail because of network problems or database permissions.
Use `fabric$status()` to inspect connection state and
`fabric$prepare("sql")` to retry preparing SQL access. A problem with one
service can leave the others available.

For application error handling, `shinyOAuth_access_error` provides a
`context$reason` field. `interaction_required` calls for an explicit sign-in
again; `insufficient_scope` calls for checking permissions and consent.
`refresh_pending` means a synchronous call cannot yet get the renewed
credential. SQL may wrap these errors in the parent of a
`fabric_sql_authentication_error`. Show users a useful message without exposing
raw credentials or provider diagnostics.

Routine token refresh does not change `generation()` or rerun queries that
depend on it. A token passed to a worker is fixed for that invocation; it does
not renew there. If it expires, start a new task to acquire a fresh token.
Synchronous calls using a token provider retain fabricQueryR's retry behavior:
one HTTP 401 response
can trigger token renewal and a request retry even with `idempotent = FALSE`.
This is separate from transient network/service retries. `prepare()` and
`reauthorize()` themselves do not replay data operations.

### Connections and deployment

`fabric_sql_query()` opens and closes its own connection. If you use
`fabric_sql_connect()` or `$sql_connect()` directly, close it with
`DBI::dbDisconnect()`, usually through `on.exit()` in a helper function. Signing
out does not close an already open DBI connection. Use `backend = "adbc"` with
its driver when exact numeric or Arrow results are needed.

Set up the future worker pool once when the app starts. The example uses two
background R processes; adjust `workers` for the resources available on your
host. Use an exact registered HTTPS callback URL for deployment.
The `uiPattern = ".*"` setting lets callback subpaths reach the app. Check your
hosting service's routing, reverse-proxy and cookie settings using shinyOAuth's
deployment guidance. The default local sign-in lifetime is eight hours; adjust
`reauth_after_seconds` for your app. App logout does not sign the visitor out of
Microsoft everywhere.

Standard Microsoft service hosts are allowed. A trusted gateway needs its exact
hostname in `endpoint_hosts`, configured by the app author. HTTP redirects are
rejected, and later request destinations are checked too. These checks protect
normal package use; they do not isolate untrusted R code.
