---
title: "Run Spark work from R with Livy"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Run Spark work from R with Livy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Apache Spark processes data using compute that runs in Fabric. **Livy** is the
service that lets an R program submit Spark code and receive its status and
output. The code runs remotely; it does not run in your local R process.

Use Spark when a transformation is too large for one computer, needs a
Spark-specific library or format, or already exists as a Spark application.
For a small table read, SQL or a direct Lakehouse reader is usually simpler and
starts faster.

## Before the first call

You need a Fabric workspace on supported capacity and a Lakehouse with a Livy
endpoint. The signed-in account must be allowed to read and execute the
Lakehouse and access the data used by the code.

Discovering the Lakehouse is normally enough; its record carries the endpoint:

```{r}
library(fabricQueryR)

workspace <- fabric_workspaces()[["Analytics workspace"]]
lakehouse <- fabric_lakehouses(workspace)[[1L]]
```

If discovery cannot retrieve the endpoint in your environment, copy the
session-job connection string from **Lakehouse settings > Livy endpoint** and
pass that URL instead.

## Run one piece of Spark code

`fabric_livy_query()` is the simplest helper. It starts a temporary session,
runs one statement, waits, and closes the session:

```{r}
result <- fabric_livy_query(
  lakehouse,
  kind = "sql",
  code = "SELECT 1 AS id, 'hello from Spark' AS message"
)

result$output$parsed
```

The returned `output$parsed` value is usually a tibble for tabular results, an
R object for JSON, or character output for printed text.

## Choose the language that matches the code

The `kind` argument tells Livy how to interpret `code`:

| `kind` | Code language |
|---|---|
| `"sql"` | Spark SQL |
| `"sparkr"` | SparkR |
| `"pyspark"` | Python with Spark |
| `"spark"` | Scala |

For example, SparkR code can use the active Spark session and Lakehouse:

```{r}
result <- fabric_livy_query(
  lakehouse,
  kind = "sparkr",
  code = paste(
    "df <- sql('SELECT * FROM orders LIMIT 100')",
    "printSchema(df)",
    "showDF(df, numRows = 10)",
    sep = "\n"
  )
)
```

The language selected by `kind` is the language running in Fabric. Local R
variables are not automatically available there; build the submitted code
deliberately and do not insert untrusted text into it.

## Reuse a session for several statements

Starting Spark can take time. Use `fabric_livy_session()` when sequential
statements need to share variables or cached data:

```{r}
session <- fabric_livy_session(lakehouse)
on.exit(session$close(), add = TRUE)

session$wait()
session$run("shared_value = 40", kind = "pyspark")
answer <- session$run("print(shared_value + 2)", kind = "pyspark")
answer$output$parsed
```

Always close a session explicitly. R object cleanup does not make a network
request, so forgetting `$close()` can leave avoidable Spark work running.

A standard session is right for one R process running a sequence. High
concurrency is an advanced option for several isolated workloads that may share
underlying compute; it is not needed for a few statements in order.

## Submit a complete application file

Use a Livy batch when the work is a repeatable Python, R, or Java/Scala script
stored in OneLake or ADLS:

```{r}
batch <- fabric_livy_batch_submit(
  lakehouse,
  file = paste0(
    "abfss://", workspace$id,
    "@onelake.dfs.fabric.microsoft.com/",
    lakehouse$id,
    ".Lakehouse/Files/jobs/daily_transform.py"
  ),
  name = "daily-transform",
  wait = TRUE,
  timeout = 1800
)

batch$result()
```

The application file must already be available through an ABFS or ABFSS path.
Upload it with `fabric_onelake_upload()` first when necessary. With
`wait = FALSE`, the function returns a `FabricLivyBatch` object immediately;
call its `$wait()`, `$result()`, or `$logs()` methods later.

## Use an Environment for repeatable configuration

A published Fabric Environment can hold Spark settings and libraries shared by
several runs. Discover it and pass its ID when the workload depends on that
configuration:

```{r}
environment <- fabric_environments(workspace)[[1L]]

result <- fabric_livy_query(
  lakehouse,
  kind = "pyspark",
  code = "print(spark.version)",
  environment_id = environment$id
)
```

Prefer an Environment over repeating a long set of configuration overrides in
every R call. Leave driver memory, executor memory, cores, and executor counts
at Fabric defaults until the workload has been measured and sized.

## Decide whether the result belongs in R

Small summaries and samples are good return values for a Livy statement. A
large transformed dataset should normally be written by Spark to a Lakehouse
table or file and then read selectively from R. This keeps network transfer and
local memory use bounded and makes the Fabric result reusable by other tools.
