---
title: "Working with Livy (Spark)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with Livy (Spark)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Apache Spark processes data using compute that runs in Fabric. *Livy* is the
service that lets an R program submit Spark code and receive its status and
output. The Spark code runs on Fabric; it does not run in your local R process.

Use Spark when a transformation is too large for one computer, needs a
Spark-specific library or format, or already exists as a Spark application.
For a small table read, SQL or a direct Lakehouse reader is usually simpler and
starts faster. This guide begins with one small statement, then moves to reusable
sessions, complete applications, and shared environments.

## Before the first call

You need a Fabric workspace on supported capacity, a Lakehouse with a Livy
endpoint, and the *tenant admin setting for the Livy API enabled*. 

A delegated caller needs all four of these Microsoft Entra scopes:
- `Lakehouse.Execute.All`
- `Lakehouse.Read.All`
- `Code.AccessFabric.All`
- `Code.AccessStorage.All`

The signed-in user must be a *Contributor* in the workspace containing the
Livy endpoint and data-source items. For unattended authentication, add the
service principal to the workspace as a Contributor. 

Then, you can discover the Lakehouse item which you want to use Livy with:

```{r, eval = FALSE}
library(fabricQueryR)

workspaces <- fabric_workspaces()
matches <- Filter(
  \(x) identical(x$displayName, "Analytics workspace"),
  workspaces
)
stopifnot(length(matches) == 1L)
workspace <- matches[[1L]]
lakehouse <- fabric_lakehouses(workspace)[[1L]]
```

If discovery cannot retrieve the endpoint in your environment, copy the
session-job connection string from *Lakehouse settings > Livy endpoint* and
pass that URL instead.

## Run one piece of Spark code

`fabric_livy_query()` is the simplest helper. It starts a temporary session,
runs one statement, waits, and closes the session:

```{r, eval = FALSE}
result <- fabric_livy_query(
  lakehouse,
  kind = "sql",
  code = "SELECT 1 AS id, 'hello from Spark' AS message"
)

result$output$parsed
```

The returned `output$parsed` value is usually a tibble for tabular results, an
R object for JSON, or character output for printed text.

## Choose the language that matches the code

The `kind` argument tells Livy how to interpret `code`:

| `kind` | Code language |
|---|---|
| `"sql"` | Spark SQL |
| `"sparkr"` | SparkR |
| `"pyspark"` | Python with Spark |
| `"spark"` | Scala |

For example, SparkR code can use the active Spark session and Lakehouse:

```{r, eval = FALSE}
result <- fabric_livy_query(
  lakehouse,
  kind = "sparkr",
  code = paste(
    "df <- sql('SELECT * FROM orders LIMIT 100')",
    "printSchema(df)",
    "showDF(df, numRows = 10)",
    sep = "\n"
  )
)
```

## Reuse a session for several statements

Starting Spark can take time. Use `fabric_livy_session()` when sequential
statements need to share variables or cached data:

```{r, eval = FALSE}
session <- fabric_livy_session(lakehouse)
on.exit(session$close(), add = TRUE)

session$wait()
session$run("shared_value = 40", kind = "pyspark")
answer <- session$run("print(shared_value + 2)", kind = "pyspark")
answer$output$parsed
```

Always close a session explicitly. R object cleanup does not make a network
request, so forgetting `$close()` can leave Spark running.

A standard session is right for one R process running a sequence. High
concurrency is an advanced option for several isolated workloads that may share
underlying compute; it is not needed for a few statements in order.

## Submit a complete application file

Use a Livy batch when the work is a repeatable Python, R, or Java/Scala script
stored in OneLake or ADLS:

```{r, eval = FALSE}
batch <- fabric_livy_batch_submit(
  lakehouse,
  file = paste0(
    "abfss://", workspace$id,
    "@onelake.dfs.fabric.microsoft.com/",
    lakehouse$id,
    ".Lakehouse/Files/jobs/daily_transform.py"
  ),
  name = "daily-transform",
  wait = TRUE,
  timeout = 1800
)

batch$result()
```

The application file must already be available through an ABFS or ABFSS path.
Upload it with `fabric_onelake_upload()` first when necessary. With
`wait = FALSE`, the function returns a `FabricLivyBatch` object immediately;
call its `$wait()`, `$result()`, or `$logs()` methods later.

If a session, statement, or batch wait times out, the
`fabric_livy_timeout_error` condition keeps the exact live object in its
`handle` field. You can inspect `$status()` or request `$cancel()` through that
handle in the current R process. The kind-specific `session`, `statement`, or
`batch` field contains safe metadata for logging; after serialization, a handle
intentionally no longer carries its credential.

## Use an Environment for repeatable configuration

A published Fabric Environment can hold Spark settings and libraries shared by
several runs. Discover it and pass its ID when the workload depends on that
configuration:

```{r, eval = FALSE}
environment <- fabric_environments(workspace)[[1L]]

result <- fabric_livy_query(
  lakehouse,
  kind = "pyspark",
  code = "print(spark.version)",
  environment_id = environment$id
)
```
