---
title: "Invoke Fabric user data functions (experimental)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Invoke Fabric user data functions (experimental)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

A Fabric *user data function* is reusable Python code that Fabric hosts and
runs. It can accept inputs, perform a task, and return a result. With
'fabricQueryR', an R session can call the published function and receive its
structured response.

## Lifecycle

The package's User Data Function discovery and invocation APIs are
experimental. Their validation, request construction, and response handling
are covered by offline tests, and lightweight Core item discovery is exercised
against Fabric. However, the package cannot currently maintain repeatable
end-to-end coverage for detailed User Data Function discovery and public
function invocation.

The persistent sandbox is provisioned with a service principal. Microsoft's
User Data Function create, update-definition, detailed Get, and delete APIs
currently support delegated users, but not service principals or managed
identities. The sandbox therefore cannot provision, publish, fully inspect, and
remove its own disposable public-function fixtures. Until repeatable end-to-end
coverage is possible, treat these APIs and their returned service shapes as
subject to change.

This guide prepares a function for external use, makes one small call, and then
introduces structured parameters and safe retries. Start with a simple function
whose output you can verify in the Fabric portal.

## Prepare the function in Fabric

In the Fabric portal, publish the user data functions item, switch to *Run
only* mode, open the function's properties, enable *Public access*, and copy
the *Public URL*. Each function has its own URL.

`fabric_user_data_functions()` and `workspace$user_data_functions()` default to
`detail = FALSE`. This uses the Core item listing, which supports delegated
users, service principals, and managed identities, and returns a read-only
`FabricItem` R6 object. `$details()` calls `fabric_item()` to refresh its item
record. To force the workload-specific Get API, use `$details(detail = TRUE)`
or rediscover with `detail = TRUE`; Microsoft documents that API for delegated
users only, not service principals or managed identities. Neither response
includes enough information to derive each public function URL, so pass the
copied URL to `fabric_function_invoke()`.

You could store the URL in an environment variable (rather than hardcoding it):

```{r, eval = FALSE}
function_url <- Sys.getenv("FABRIC_FUNCTION_URL")
```

The caller needs Execute permission on the item. A delegated sign-in also needs
one Power BI delegated permission. By default, 'fabricQueryR' requests the
least-privilege `UserDataFunction.Execute.All` scope. Microsoft also accepts
the broader `Item.Execute.All` scope, but it does not replace the caller's item
Execute permission. If the app registration grants only that broader scope,
select it explicitly:

```{r, eval = FALSE}
result <- fabric_function_invoke(
  function_url,
  parameters = list(customerName = "Ada", priority = 2L),
  audience = paste0(
    "https://analysis.windows.net/powerbi/api/",
    "Item.Execute.All"
  )
)
```

If sign-in succeeds but the call is denied, ask the item owner or Fabric
administrator to check item Execute access and the relevant tenant setting.

## Make a first call

Parameter names must match the published Python function.
For a function with `customerName` and `priority` inputs:

```{r, eval = FALSE}
result <- fabric_function_invoke(
  function_url,
  parameters = list(
    customerName = "Ada",
    priority = 2L
  )
)

result$status
result$output
```

## Send structured parameters

The request body is one named JSON object. Its names must match the published
function's camelCase parameters:

```{r, eval = FALSE}
structured_result <- fabric_function_invoke(
  function_url,
  parameters = list(
    customerName = "Ada",
    priority = 2L,
    lineIds = I(c(101L, 102L)),
    metadata = list(source = "R", approved = TRUE, note = NULL)
  )
)
```

One-element R values normally become JSON scalars. Wrap a one-element value in
`I()` when the Python signature expects a list. Named atomic vectors and data
frames are also accepted as top-level parameter objects; a named list gives the
clearest control over nested JSON shapes.

The result is a list which can be inspected:

```{r, eval = FALSE}
structured_result$function_name
structured_result$invocation_id
structured_result$status
structured_result$output
structured_result$errors
structured_result$http_status
```

The response keeps the function status, output, and any structured errors
together. Inspect these fields before using the output in a later step.

## Retry only safe functions

A function can write data or call another service, so invocation POSTs are not
retried by default. If a function is read-only or implements its own durable
idempotency key, opt in explicitly:

```{r, eval = FALSE}
result <- fabric_function_invoke(
  function_url,
  parameters = list(requestId = "stable-business-key"),
  idempotent = TRUE
)
```

Opted-in calls use bounded retries for transport failures, throttling, and
transient HTTP responses.

## Limits

Public function calls have request-size, execution-time, and response-size
limits. Keep calls focused and pass large data through a Fabric data store
instead of function parameters. See the service-limits link below for current
values.

## More information

See Microsoft's [external invocation
tutorial](https://learn.microsoft.com/en-us/fabric/data-engineering/user-data-functions/tutorial-invoke-from-python-app),
[programming model](https://learn.microsoft.com/en-us/fabric/data-engineering/user-data-functions/python-programming-model),
and [service limits](https://learn.microsoft.com/en-us/fabric/data-engineering/user-data-functions/user-data-functions-service-limits).
