---
title: "Invoke Fabric user data functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Invoke Fabric user data functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

A Fabric user data function is similar to calling a function in another R
package, except its Python code runs in Fabric. R sends named inputs to a secure
web address and receives a structured result.

fabricQueryR invokes functions that are already defined, published, and enabled
for public access. It does not create or deploy the function. This guide starts
with one small scalar call before introducing nested inputs, retries, and
security limits.

## Prepare the function in Fabric

In the Fabric portal, publish the user data functions item, switch to **Run
only** mode, open the function's properties, enable **Public access**, and copy
the **Public URL**. Each function has its own URL. Although
`fabric_user_data_functions()` discovers the containing item, Microsoft's item
response does not expose enough information to derive each public URL safely.

Store the copied URL in configuration rather than source code:

```{r, eval = FALSE}
function_url <- Sys.getenv("FABRIC_FUNCTION_URL")
```

The caller needs Execute permission on the item. Interactive app registrations
need the Power BI delegated permission `UserDataFunction.Execute.All` or
`Item.Execute.All`. Application credentials require the relevant
service-principal tenant setting and item access. fabricQueryR chooses the
narrow delegated scope for interactive authentication and the Power BI
`.default` audience for client credentials; use `audience` only for a custom
identity provider.

The caller identity for the public endpoint and the connections used inside the
function are separate concerns. Microsoft currently does not support using a
service principal through UDF-managed connections to access Fabric items or
data sources. An application can therefore be allowed to invoke a function
whose own managed data connection remains unsupported for that identity.

## Make a first call

Parameter names must match the published Python function. For a function with
`customerName` and `priority` inputs:

```{r, eval = FALSE}
result <- fabric_function_invoke(
  function_url,
  parameters = list(
    customerName = "Ada",
    priority = 2L
  )
)

result$status
result$output
```

`Succeeded` means the function completed. The output shape is defined by the
published function, so inspect `result$output` before building downstream R
code around it.

## Send structured parameters

The request body is one named JSON object. Its names must match the published
function's camelCase parameters:

```{r, eval = FALSE}
structured_result <- fabric_function_invoke(
  function_url,
  parameters = list(
    customerName = "Ada",
    priority = 2L,
    lineIds = I(c(101L, 102L)),
    metadata = list(source = "R", approved = TRUE, note = NULL)
  )
)
```

One-element R values normally become JSON scalars. Wrap a one-element value in
`I()` when the Python signature expects a list. Named atomic vectors and data
frames are also accepted as top-level parameter objects; a named list gives the
clearest control over nested JSON shapes.

The result is deliberately inspectable:

```{r, eval = FALSE}
structured_result$function_name
structured_result$invocation_id
structured_result$status
structured_result$output
structured_result$errors
structured_result$http_status
```

Fabric documents the statuses `Succeeded`, `BadRequest`, `Failed`, `Timeout`,
and `ResponseTooLarge`. A valid invocation envelope is returned even when the
corresponding HTTP response is 400, 403, 408, 409, 422, or 500. This keeps a
handled `UserThrownError`, its message, and its `properties` available to the R
application. Authentication, item authorization, throttling, malformed JSON,
and non-function service responses still raise typed package errors.

The invocation tutorial describes an error `name`; current live responses can
instead provide `errorCode`. `result$errors` adds `name` as an alias when only
`errorCode` is present, while `result$response` retains the original service
shape.

Unknown response fields and future status values remain in `result$response`.
Integers outside R's exact double range are preserved as character values.

## Retry only safe functions

A function can write data or call another service, so invocation POSTs are not
retried by default. If a function is read-only or implements its own durable
idempotency key, opt in explicitly:

```{r, eval = FALSE}
result <- fabric_function_invoke(
  function_url,
  parameters = list(requestId = "stable-business-key"),
  idempotent = TRUE
)
```

Opted-in calls use bounded retries for transport failures, throttling, and
transient HTTP responses. `idempotent = TRUE` is a promise made by the caller;
fabricQueryR cannot infer whether repeating arbitrary Python code is safe.

## Trust boundary, limits, and redaction

Bearer tokens are sent to the exact `function_url` supplied by the caller after
the URL passes structural validation. Supplying that URL is the explicit trust
decision, so verify custom origins independently. URLs must use HTTPS and the
documented `/v1/workspaces/.../userDataFunctions/.../functions/.../invoke`
route. Embedded credentials, query strings, fragments, and nonstandard ports
are rejected.

Microsoft currently limits the combined request parameters to 4 MB, execution
through a public endpoint to 100 seconds, and the function return value to 30
MB. fabricQueryR rejects oversized requests before authentication, waits up to
110 seconds by default so Fabric can return its own timeout envelope, and caps
the complete response at 32 MiB to leave room around a valid 30 MB output.

Returned values and errors can otherwise become logs or persisted objects.
fabricQueryR redacts secret-named fields such as `token`, `authorization`,
`password`, and `apiKey`, as well as bearer-token text, from errors, response
metadata, and attached conditions. Function `output` is returned unchanged
because those names can be legitimate domain data. Treat output as untrusted
application data and do not use a user data function as a secret-delivery
mechanism.

For the current service contract, see Microsoft's [external invocation
tutorial](https://learn.microsoft.com/en-us/fabric/data-engineering/user-data-functions/tutorial-invoke-from-python-app),
[programming model](https://learn.microsoft.com/en-us/fabric/data-engineering/user-data-functions/python-programming-model),
and [service limits](https://learn.microsoft.com/en-us/fabric/data-engineering/user-data-functions/user-data-functions-service-limits).
