---
title: "Invoke Fabric user data functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Invoke Fabric user data functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

A Fabric *user data function* is a reusable piece of Python code that runs in
Microsoft Fabric. It can accept inputs, perform a task, and return a result.
Fabric hosts and runs the Python code.

With fabricQueryR, you can invoke these functions from R. R then sends named
inputs to the function's secure web address, Fabric runs the function, and R
receives a structured response. This vignette show a simple example.

## Prepare the function in Fabric

In the Fabric portal, publish the user data functions item, switch to *Run
only* mode, open the function's properties, enable *Public access*, and copy
the *Public URL*. Each function has its own URL. Although
`fabric_user_data_functions()` discovers the containing item, Microsoft's item
response does not expose enough information to derive each public URL.

You could store the URL in an environment variable (rather than hardcoding it):

```{r, eval = FALSE}
function_url <- Sys.getenv("FABRIC_FUNCTION_URL")
```

The function caller needs Execute permission on the item. Interactive app registrations
need the Power BI delegated permission `UserDataFunction.Execute.All` or
`Item.Execute.All`. Application credentials require the relevant
service-principal tenant setting and item access.

## Make a first call

Parameter names must match the published Python function. 
For a function with `customerName` and `priority` inputs:

```{r, eval = FALSE}
result <- fabric_function_invoke(
  function_url,
  parameters = list(
    customerName = "Ada",
    priority = 2L
  )
)

result$status
result$output
```

## Send structured parameters

The request body is one named JSON object. Its names must match the published
function's camelCase parameters:

```{r, eval = FALSE}
structured_result <- fabric_function_invoke(
  function_url,
  parameters = list(
    customerName = "Ada",
    priority = 2L,
    lineIds = I(c(101L, 102L)),
    metadata = list(source = "R", approved = TRUE, note = NULL)
  )
)
```

One-element R values normally become JSON scalars. Wrap a one-element value in
`I()` when the Python signature expects a list. Named atomic vectors and data
frames are also accepted as top-level parameter objects; a named list gives the
clearest control over nested JSON shapes.

The result is a list which can be inspected:

```{r, eval = FALSE}
structured_result$function_name
structured_result$invocation_id
structured_result$status
structured_result$output
structured_result$errors
structured_result$http_status
```

Fabric documents the statuses `Succeeded`, `BadRequest`, `Failed`, `Timeout`,
and `ResponseTooLarge`. A valid invocation envelope is returned even when the
corresponding HTTP response is 400, 403, 408, 409, 422, or 500. This keeps a
handled `UserThrownError`, its message, and its `properties` available.

## Retry only safe functions

A function can write data or call another service, so invocation POSTs are not
retried by default. If a function is read-only or implements its own durable
idempotency key, opt in explicitly:

```{r, eval = FALSE}
result <- fabric_function_invoke(
  function_url,
  parameters = list(requestId = "stable-business-key"),
  idempotent = TRUE
)
```

Opted-in calls use bounded retries for transport failures, throttling, and
transient HTTP responses. 

## Limits

Microsoft Fabric currently limits the combined request parameters to 4 MB, execution
through a public endpoint to 100 seconds, and the function return value to 30
MB. fabricQueryR rejects oversized requests before authentication, waits up to
110 seconds by default so Fabric can return its own timeout envelope, and caps
the complete response at 32 MiB to leave room around a valid 30 MB output.

## More information

See Microsoft's [external invocation
tutorial](https://learn.microsoft.com/en-us/fabric/data-engineering/user-data-functions/tutorial-invoke-from-python-app),
[programming model](https://learn.microsoft.com/en-us/fabric/data-engineering/user-data-functions/python-programming-model),
and [service limits](https://learn.microsoft.com/en-us/fabric/data-engineering/user-data-functions/user-data-functions-service-limits).
