---
title: "Write R and Arrow data to Fabric Warehouse"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Write R and Arrow data to Fabric Warehouse}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

A Fabric Warehouse stores relational tables that you can query with SQL. Use
`fabric_warehouse_write_table()` when rows held in R should become part of one
of those tables. The source can be an ordinary data frame or, later, a larger
Arrow dataset.

This guide starts by appending a three-row data frame. It then covers table
creation, large inputs, replacement choices, and the staging and transaction
details that matter in production.

## Find the Warehouse and a staging Lakehouse

A Warehouse is the final destination. The writer also needs a Lakehouse in the
same tenant for temporary files while Fabric loads the data. Those files are
removed after a confirmed successful write.

Discover both items so their IDs and connection details travel with them:

```{r}
library(fabricQueryR)

workspace <- fabric_workspaces()[["Analytics"]]
warehouse <- fabric_warehouses(workspace)[[1L]]
staging_lakehouse <- fabric_lakehouses(workspace)[[1L]]
```

## Append a small R data frame

Start with a test table whose definition already matches the R columns:

```{r}
written <- fabric_warehouse_write_table(
  warehouse,
  table = "orders",
  data = data.frame(
    id = 1:3,
    label = c("alpha", "beta", "gamma"),
    amount = c(10.5, NA, 30)
  ),
  staging_lakehouse = staging_lakehouse,
  schema = "dbo",
  mode = "Append"
)

written$rows
written$file_count
written$staging_retained
```

`rows` confirms how many source rows were staged. A successful result normally
has `staging_retained = FALSE`, meaning the temporary OneLake files were
removed.

Column identifiers are quoted and mapped in the same ordinal order as the
Parquet schema. The names and compatible types therefore need to match the
existing destination columns. Factor columns are serialized as strings; Arrow
retains supported 64-bit integers, dates, timestamps, decimals, and nulls.

To create a missing target and load it in the same transaction:

```{r}
created <- fabric_warehouse_write_table(
  warehouse,
  table = "orders_from_r",
  data = orders,
  staging_lakehouse = staging_lakehouse,
  create_if_missing = TRUE
)
```

Fabric infers the names and Warehouse types from the staged Parquet result.
Use a pre-created table when exact lengths, nullability, constraints, or other
schema details must be controlled explicitly.

The SQL identity needs `INSERT` and `ADMINISTER DATABASE BULK OPERATIONS` on
the Warehouse. Microsoft also documents Contributor or higher workspace access
for the OneLake staging source and Warehouse destination. Cross-workspace
staging is supported in the same tenant through `staging_workspace` when the
Lakehouse is supplied by name or ID.

Behind the scenes, the function writes bounded Parquet parts to OneLake and
uses Fabric's high-throughput `COPY INTO` path. When creating a missing table,
it uses a create-table-as-select operation so creation and loading are one
transaction.

## Stream a larger-than-memory Arrow source

Arrow Datasets, Scanners, dplyr queries, RecordBatchReaders, Tables, and
Arrow-compatible streams use the same function. They are consumed one record
batch at a time rather than collected into an R data frame:

```{r}
dataset <- arrow::open_dataset("local-parquet-directory")

written <- fabric_warehouse_write_table(
  warehouse,
  table = "orders",
  data = dataset,
  staging_lakehouse = staging_lakehouse,
  target_file_size = 512 * 1024^2,
  max_rows_per_file = NULL
)
```

`target_file_size` is a soft compressed-size boundary. The 512 MB default is
inside Fabric's documented 100 MB to 1 GB performance range. Use
`max_rows_per_file` when an exact and deterministic part boundary matters.

## Choose truncate or drop replacement

```{r}
replaced <- fabric_warehouse_write_table(
  warehouse,
  table = "orders",
  data = replacement,
  staging_lakehouse = staging_lakehouse,
  mode = "Overwrite",
  overwrite_method = "Truncate"
)
```

The default truncate path preserves the existing definition, executes
`COPY INTO`, and commits only after the load succeeds. Use drop replacement
when the target definition should instead be inferred again:

```{r}
recreated <- fabric_warehouse_write_table(
  warehouse,
  table = "orders",
  data = replacement,
  staging_lakehouse = staging_lakehouse,
  mode = "Overwrite",
  overwrite_method = "Drop",
  create_if_missing = TRUE
)
```

Drop replacement discards the old table definition—including table-specific
constraints, indexes, and grants—and recreates it with CTAS. Setting
`create_if_missing = TRUE` also lets this path proceed when the target is
already absent. Fabric Warehouse documents DDL inside ACID transactions and
all-or-nothing rollback for these grouped write operations.

Local temporary files are always removed. The OneLake staging directory is
removed only after confirmed success. If SQL delivery becomes ambiguous, it is
retained even when `keep_staging_on_failure = FALSE`; the error condition
contains `staging_path`, `staging_retained`, and `ambiguous` fields for recovery
and diagnosis.

## Microsoft documentation

- [COPY INTO in Fabric Warehouse](https://learn.microsoft.com/en-us/sql/t-sql/statements/copy-into-transact-sql?view=fabric)
- [Create tables in Fabric Warehouse](https://learn.microsoft.com/en-us/fabric/data-warehouse/create-table)
- [Query Parquet files](https://learn.microsoft.com/en-us/fabric/data-warehouse/query-parquet-files)
- [Warehouse performance guidelines](https://learn.microsoft.com/en-us/fabric/data-warehouse/guidelines-warehouse-performance)
- [Warehouse transactions](https://learn.microsoft.com/en-us/fabric/data-warehouse/transactions)
