---
title: "Working with Fabric Warehouses"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with Fabric Warehouses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

A Fabric Warehouse stores data in relational tables, much like a traditional
SQL database. You normally use its SQL endpoint to read and query those tables.
This guide first connects to a Warehouse, then shows how to add or replace data
from R.

## Find and connect to a Warehouse

Start by finding the workspace and Warehouse by name. The returned `warehouse`
object contains the IDs and SQL connection details that the other functions
need:

```{r, eval = FALSE}
library(fabricQueryR)

workspaces <- fabric_workspaces()
matches <- Filter(\(x) identical(x$displayName, "Analytics"), workspaces)
stopifnot(length(matches) == 1L)
workspace <- matches[[1L]]
warehouse <- fabric_warehouses(workspace)[[1L]]
```

The SQL endpoint is the address that database tools use to reach the
Warehouse. Because it is included in `warehouse`, you do not need to find or
copy that address from the Fabric portal.

For a single query, pass that object directly to `fabric_sql_query()`:

```{r, eval = FALSE}
orders <- fabric_sql_query(
  warehouse,
  "SELECT TOP 10 * FROM dbo.orders"
)
```

The function opens and closes the SQL connection for you. If you want to run
several commands with 'DBI', open a reusable connection instead:

```{r, eval = FALSE}
con <- fabric_sql_connect(warehouse)
DBI::dbListTables(con)
DBI::dbGetQuery(con, "SELECT TOP 10 * FROM dbo.orders")
DBI::dbDisconnect(con)
```

For a simple read, `fabric_warehouse_read_table()` lets you name a table and
optionally select columns or limit the rows, without writing SQL. It uses SQL
internally. `fabric_sql_read_table()` is the more general version for any
supported Fabric SQL item.

Use `fabric_sql_query()` when you need filters, joins, grouping, or other SQL.
Use `fabric_sql_connect()` when you want to keep a connection open for several
'DBI' calls. To add or replace many rows from an R data frame or Arrow source,
use `fabric_warehouse_write_table()`. See `vignette("reading-data")` for more
reading examples.

## Prepare a staging Lakehouse for writes

Writing to a Warehouse also requires a Lakehouse in the same tenant. The
package temporarily stores files there while Fabric loads them, then removes
the files after a confirmed successful write:

```{r, eval = FALSE}
staging_lakehouse <- fabric_lakehouses(workspace)[[1L]]
```

## Add rows to an existing table

Suppose `dbo.orders` already contains order data. This call adds three new rows
to the table; it does not remove or change the rows already there. The existing
table columns must match the R data frame:

```{r, eval = FALSE}
written <- fabric_warehouse_write_table(
  warehouse,
  table = "orders",
  data = data.frame(
    id = 1:3,
    label = c("alpha", "beta", "gamma"),
    amount = c(10.5, NA, 30)
  ),
  staging_lakehouse = staging_lakehouse,
  schema = "dbo",
  mode = "Append"
)

written$rows
written$file_count
written$staging_retained
```

If you want to create a new table, you can use `create_if_missing = TRUE`:

```{r, eval = FALSE}
created <- fabric_warehouse_write_table(
  warehouse,
  table = "orders_from_r",
  data = orders,
  staging_lakehouse = staging_lakehouse,
  create_if_missing = TRUE
)
```

## Stream a larger-than-memory Arrow source

Arrow Datasets, Scanners, 'dplyr' queries, RecordBatchReaders, Tables, and
Arrow-compatible streams use the same function. They are consumed one record
batch at a time rather than collected into an R data frame:

```{r, eval = FALSE}
dataset <- arrow::open_dataset("local-parquet-directory")

written <- fabric_warehouse_write_table(
  warehouse,
  table = "orders",
  data = dataset,
  staging_lakehouse = staging_lakehouse,
  target_file_size = 512 * 1024^2,
  max_rows_per_file = NULL
)
```

`target_file_size` is a soft compressed-size boundary. The 512 MB default is
inside Fabric's documented 100 MB to 1 GB performance range. Use
`max_rows_per_file` when an exact and deterministic part boundary matters.

## Choose truncate or drop replacement

```{r, eval = FALSE}
replaced <- fabric_warehouse_write_table(
  warehouse,
  table = "orders",
  data = replacement,
  staging_lakehouse = staging_lakehouse,
  mode = "Overwrite",
  overwrite_method = "Truncate"
)
```

The default truncate path preserves the existing definition, executes
`COPY INTO`, and commits only after the load succeeds. Use drop replacement
when the target definition should instead be inferred again:

```{r, eval = FALSE}
recreated <- fabric_warehouse_write_table(
  warehouse,
  table = "orders",
  data = replacement,
  staging_lakehouse = staging_lakehouse,
  mode = "Overwrite",
  overwrite_method = "Drop",
  create_if_missing = TRUE
)
```

Drop replacement discards the old table definition—including table-specific
constraints, indexes, and grants—and recreates it with CTAS. Setting
`create_if_missing = TRUE` also lets this path proceed when the target is
already absent. Fabric Warehouse documents DDL inside ACID transactions and
all-or-nothing rollback for these grouped write operations.

Local temporary files are always removed. The OneLake staging directory is
removed only after confirmed success. If SQL delivery becomes ambiguous, it is
retained even when `keep_staging_on_failure = FALSE`; the error condition
contains `staging_path`, `staging_retained`, and `ambiguous` fields for recovery
and diagnosis.

## Microsoft documentation

- [COPY INTO in Fabric Warehouse](https://learn.microsoft.com/en-us/sql/t-sql/statements/copy-into-transact-sql?view=fabric)
- [Create tables in Fabric Warehouse](https://learn.microsoft.com/en-us/fabric/data-warehouse/create-table)
- [Query Parquet files](https://learn.microsoft.com/en-us/fabric/data-warehouse/query-parquet-files)
- [Warehouse performance guidelines](https://learn.microsoft.com/en-us/fabric/data-warehouse/guidelines-warehouse-performance)
- [Warehouse transactions](https://learn.microsoft.com/en-us/fabric/data-warehouse/transactions)
