---
title: "Working with Fabric Warehouses"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with Fabric Warehouses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

A Fabric Warehouse stores data in relational tables, much like a traditional
SQL database. You normally use its SQL endpoint to read and query those tables.
This guide first connects to a Warehouse, then shows how to add or replace data
from R.

## Find and connect to a Warehouse

Start by finding the workspace and Warehouse by name. The returned `warehouse`
is a read-only `FabricWarehouse` R6 object. It keeps its Fabric fields, IDs,
SQL connection details, and credential, and provides methods for the useful
next actions:

```{r, eval = FALSE}
library(fabricQueryR)

workspaces <- fabric_workspaces()
matches <- Filter(\(x) identical(x$displayName, "Analytics"), workspaces)
stopifnot(length(matches) == 1L)
workspace <- matches[[1L]]
warehouse <- workspace$warehouses()[[1L]]
```

The SQL endpoint is the address that database tools use to reach the
Warehouse. Because it is included in `warehouse`, you do not need to find or
copy that address from the Fabric portal.

For a single query, call the object's `$sql_query()` method:

```{r, eval = FALSE}
orders <- warehouse$sql_query(
  "SELECT TOP 10 * FROM dbo.orders"
)
```

The function opens and closes the SQL connection for you. If you want to run
several commands with 'DBI', open a reusable connection instead:

```{r, eval = FALSE}
con <- warehouse$sql_connect()
DBI::dbListTables(con)
DBI::dbGetQuery(con, "SELECT TOP 10 * FROM dbo.orders")
DBI::dbDisconnect(con)
```

For a simple read, `$read_table()` lets you name a table and
optionally select columns or limit the rows, without writing SQL. It uses SQL
internally. `fabric_sql_read_table()` is the more general version for any
supported Fabric SQL item.

Use `$sql_query()` when you need filters, joins, grouping, or other SQL.
Use `$sql_connect()` when you want to keep a connection open for several
'DBI' calls. To add or replace many rows from an R data frame or Arrow source,
use `$write_table()`. See `vignette("reading-data")` for more
reading examples.

## Prepare a staging Lakehouse for writes

Writing to a Warehouse also requires a Lakehouse in the same tenant. The
package temporarily stores files there while Fabric loads them, then removes
the files after a confirmed successful write:

```{r, eval = FALSE}
staging_lakehouse <- workspace$lakehouses()[[1L]]
```

## Add rows to an existing table

Suppose `dbo.orders` already contains order data. This call adds three new rows
to the table; it does not remove or change the rows already there. The existing
table columns must match the R data frame:

```{r, eval = FALSE}
written <- warehouse$write_table(
  table = "orders",
  data = data.frame(
    id = 1:3,
    label = c("alpha", "beta", "gamma"),
    amount = c(10.5, NA, 30)
  ),
  staging_lakehouse = staging_lakehouse,
  schema = "dbo",
  mode = "Append"
)

written$rows
written$file_count
written$staging_retained
```

If you want to create a new table, you can use `create_if_missing = TRUE`:

```{r, eval = FALSE}
created <- warehouse$write_table(
  table = "orders_from_r",
  data = orders,
  staging_lakehouse = staging_lakehouse,
  create_if_missing = TRUE
)
```

## Replace table data

Use overwrite mode when the new data should replace the current rows:

```{r, eval = FALSE}
replaced <- warehouse$write_table(
  table = "orders",
  data = replacement,
  staging_lakehouse = staging_lakehouse,
  mode = "Overwrite",
  overwrite_method = "Truncate"
)
```

The default `"Truncate"` method preserves the existing table definition. Use
`"Drop"` only when Fabric should infer a new definition from the incoming data:

```{r, eval = FALSE}
recreated <- warehouse$write_table(
  table = "orders",
  data = replacement,
  staging_lakehouse = staging_lakehouse,
  mode = "Overwrite",
  overwrite_method = "Drop",
  create_if_missing = TRUE
)
```

Dropping a table also removes its table-specific constraints, indexes, and
grants. Prefer `"Truncate"` unless recreating the definition is intentional.
The writer removes staging files after confirmed success and reports a retained
staging path when manual recovery may be needed.

## Scale up with Arrow

The same writer accepts Arrow Datasets, Scanners, 'dplyr' queries,
RecordBatchReaders, Tables, and Arrow-compatible streams. These sources are
processed in batches instead of first being collected into an R data frame:

```{r, eval = FALSE}
dataset <- arrow::open_dataset("local-parquet-directory")

written <- warehouse$write_table(
  table = "orders",
  data = dataset,
  staging_lakehouse = staging_lakehouse
)
```

See `?fabric_warehouse_write_table` when you need to tune staged file sizes or
recover a retained staging directory after a failed write.

## More information

- [COPY INTO in Fabric Warehouse](https://learn.microsoft.com/en-us/sql/t-sql/statements/copy-into-transact-sql?view=fabric)
- [Create tables in Fabric Warehouse](https://learn.microsoft.com/en-us/fabric/data-warehouse/create-table)
- [Query Parquet files](https://learn.microsoft.com/en-us/fabric/data-warehouse/query-parquet-files)
- [Warehouse performance guidelines](https://learn.microsoft.com/en-us/fabric/data-warehouse/guidelines-warehouse-performance)
- [Warehouse transactions](https://learn.microsoft.com/en-us/fabric/data-warehouse/transactions)
